Verified translation has two well-studied extremes: prove the translator
once (certified compilation), or validate each run of one translator
(translation validation). Both treat a single translation in isolation.
We study what happens when translations form a \emph{graph}: many source
languages, several reasoning targets, and multiple independently built
routes between them --- and when the honest answer to ``is this
translation correct?'' differs from edge to edge.

We present a calculus of \emph{fidelity-graded translations}. Its unit is
the \emph{pair}: a deterministic translator between two formally defined
languages, packaged with interpreters for both and a
\emph{target-to-source} interpreter that carries target-side behaviors
(solver witnesses, traces) back to source-side behaviors --- a commuting
square whose closure is a decidable, per-program check. Squares compose
by a pasting theorem whose hypothesis, a projection-compatibility
condition, makes explicit what each hop is allowed to discard. Every edge
declares a fidelity grade --- predicted, reproducible, checked, proved,
trusted --- and we show grades compose by weakest link on an assurance
lattice, can be \emph{re-established} per run by inline checking, and are
\emph{exceeded} by agreement between independently derived routes. A
trusted-base ledger makes each cross-check's contribution exact, and an
end-to-end theorem isolates a fundamental asymmetry: answers carrying
witnesses are self-certifying at the source after carry-back and replay,
while universal answers are precisely where grades, branches, and
certificates earn their cost. The compositional core is mechanized in
Lean~4, most of it axiom-free.

The calculus is implemented in a platform of 13 registered languages and
13 pairs --- ISAs (RISC-V, AArch64, eBPF, Wasm, EVM), C, Python, chemical
reaction networks, and molecular graphs, converging on two reasoning hubs
(BTOR2, SMT-LIB) --- built as a \emph{two-directional experiment in
LLM-generated correctness}: every pair was written by an independent LLM
agent, largely unsupervised, with the squares, differentials, and branch
cross-checks as the only semantic gate; and the platform's intended
\emph{player} is itself an LLM, handed deterministic, graded, checked
moves along paths of pairs. All code, and most of this paper, is
LLM-generated; the human contribution is the architecture. We report a
capability snapshot: per-construct coverage conjoined with fidelity for
every pair, branch-agreement experiments along dual RISC-V and AArch64
routes down to SMT-LIB, end-to-end case studies with source-level witness
replay, a certified-unreachability run whose DRAT certificate is
re-validated by an independent checker, and the defects the architecture
caught --- including one in that checker's own adapter, exposed by the
negative control.
