%% The arXiv abstract — the whole story, short: the calculus, the
%% platform, the two planes, the vision; details rely on the paper.
%% (Formerly a fork of the retired POPL variant; tags popl27-*.)

To answer a question about a program, move the program to where the
question is decidable. Every such move is a translation, and every
translation is a place to be wrong. We study translation as a
\emph{graph} --- many languages, a few reasoning targets, independently
built routes of honestly different trustworthiness --- and give it a
calculus: pairs of languages close commuting squares that are
\emph{directional} (exactness is the identity-embedding special case of
over-approximation), checkable per program, and composable, a route's
contract being the componentwise meet of its hops' --- assurance class,
direction, kept observables, measured cost. One asymmetry organizes
trust: witness-carrying answers are self-certifying by replay at the
source; universal answers are where grades, independent branches, and
re-checked certificates earn their cost. The compositional core, lax
telescope included, is mechanized in Lean~4.

hurdy-gurdy implements the calculus as \emph{two planes} meeting in one
registry. The \emph{use} plane reads declarations and produces
evidence-carrying answers; its builders and its intended player are
both LLMs, untrusted by construction. The \emph{evolution} plane grows
the graph: unmet questions are recorded as demand, pairs are
\emph{recommended by evidence} and registered by humans, and a ratchet
keeps every prior verdict standing. \emph{Answers never write; growth
never answers.} Run indefinitely, the loop converges on every reducibly
decidable question, at fidelity that only rises. We measure the July
2026 snapshot --- per-construct conjoined coverage, dual-route branch
agreement for two ISAs, source-level witness replay, certified
unreachability re-validated by a formally verified checker, escape
rates for the gate itself --- and report the defects the architecture
caught in its own authors' work.
