%% ARXIV-ONLY fork of sections/calculus.tex — arxiv.tex inputs this file;
%% main.tex keeps sections/calculus.tex untouched (the POPL submission is
%% frozen). In this version the DIRECTIONAL square is the primary notion
%% and the exact square is its identity-embedding special case; the POPL
%% version presents the exact square only, and its former \ifarxiv
%% directional subsection is absorbed here. Cherry-pick toward a future
%% POPL revision by diffing against sections/calculus.tex.
%%
%% NUMBERING (v3): §3 keeps 3.1–3.11 (3.10 = prop:exactid, 3.11 =
%% prop:lax). Since the v3 two-plane restructure the §4/§5 numbering
%% diverges from published v1/v2 (coverage/ratchet live in §5, the
%% economy; the former §4 is appended here as 3.12+, so the end-to-end
%% theorems are 3.17/3.18) — body-arxiv.tex
%% titles track this; the crosswalk guard binds main.pdf only.

\section{The machine: a calculus of pairs and its guarantees}
\label{sec:calculus}

This section is the whole use plane, in the introduction's order:
languages, behaviors, and projections (\S\ref{sec:languages}); the
directional commuting square (\S\ref{sec:pairs}); composition ---
pasting, and the contract meet (\S\ref{sec:pasting},
\S\ref{sec:grades}); the direction axis at work (\S\ref{sec:lax});
trust beyond the meet --- re-establishment, branches, and the
trusted-base ledger (\S\ref{sec:composition},
\S\ref{sec:branching}); the end-to-end asymmetry
(\S\ref{sec:endtoend}); and the mechanization (\S\ref{sec:mech}).
Coverage and the ratchet are evolution-plane machinery and live with
the economy (\S\ref{sec:coverage}).

\subsection{Languages, behaviors, projections}
\label{sec:languages}

\begin{definition}[Language]
\label{def:language}
A \emph{language} $A$ is a triple $(\Prog{A}, F_A, \sem{\cdot}{A})$ where
$\Prog{A}$ is a decidable set of \emph{programs}; $F_A$ is a finite set of
named \emph{observable fields}, each $f \in F_A$ with a value domain
$V_f$; and $\sem{\cdot}{A} : \Prog{A} \rightharpoonup \Beh{A}$ is the
\emph{reference semantics}, a partial function into behaviors.
An \emph{observable state} is a total map
$\sigma \in \Obs{A}$, where $\Obs{A} = \prod_{f \in F_A} V_f$; a
\emph{behavior} is a finite sequence of observable states,
$\Beh{A} = \Obs{A}^{*}$, recorded \emph{post-step}: the $i$-th state is
the observable state after the $i$-th transition.
\end{definition}

The only admission criterion for a language is that $\sem{\cdot}{A}$ is
\emph{definable} --- an ISA manual, a logic's model theory, a reaction-network
semantics, and a molecular-graph notation all qualify. Nothing in this
section assumes languages are executable or even computational.

The scope, stated plainly: behaviors are \emph{finite} and the
reference semantics is a \emph{function} --- nondeterminism,
divergence, and reactive I/O are outside. Behavior \emph{inclusion} is
inside in exactly one controlled form: the square is a
\emph{directional} claim (\Cref{def:pair,def:faithful}), checked as a
projected trace equality along a declared witness embedding —
two-sided ($\mathsf{exact}$) or the simulation half
($\mathsf{over}$); nothing else is admitted. This is the right notion
for near-structure-preserving translation and the wrong one for
optimizing compilation, whose home is full simulation over retimed
structure --- accordingly the one optimizing hop in the instantiation
carries no square at all, and C's non-functional official semantics
enters through a determinized fragment, policed differentially
(\Cref{sec:instantiation}).

A program here is a \emph{closed} configuration: free inputs, when present,
are part of $p$ (a program together with an input valuation). Open programs
--- programs quantified over their inputs --- enter the square through
their \emph{closing valuations} (\Cref{def:faithful} quantifies over
them) and take center stage in \S\ref{sec:endtoend}, where a solver
decides over all inputs and a witness closes one.

\begin{definition}[Interpreter]
\label{def:interpreter}
An \emph{interpreter} for $A$ is a computable partial function
$I_A : \Prog{A} \rightharpoonup \Beh{A}$ that is \emph{pure}: identical
input yields byte-identical output, on every host (the contract;
\S\ref{sec:eval-perf} enforces it by twice-and-diff on the evaluation
host). Outside its domain
it fails with a typed $\unsupported$ verdict naming the construct;
$\dom(I_A)$ is $A$'s \emph{covered fragment}.
\end{definition}

\begin{assumption}[Interpreter adequacy]
\label{asm:adequacy}
$I_A(p) = \sem{p}{A}$ for all $p \in \dom(I_A)$.
\end{assumption}

Adequacy is an assumption, not a theorem, and we keep it visible: the
irreducible residue in every trusted-base computation of
\S\ref{sec:tcb}, discharged \emph{empirically} by differential testing
against independently produced oracles (\Cref{sec:instantiation};
Python is the honest exception --- its pinned runtime is both semantics
and interpreter).

\begin{definition}[Projection]
\label{def:projection}
A \emph{projection} over $A$ is a subset $\proj \subseteq F_A$. It lifts
to states by restriction, $\proj(\sigma) = \sigma\!\restriction_{\proj}$,
and to behaviors pointwise,
$\proj(\sigma_1 \cdots \sigma_n) = \proj(\sigma_1) \cdots \proj(\sigma_n)$
(in particular, projected behaviors of different lengths are unequal).
Write $b \eqproj{\proj} b'$ for $\proj(b) = \proj(b')$.
\end{definition}

\subsection{Pairs and the directional square}
\label{sec:pairs}

\begin{definition}[Pair]
\label{def:pair}
A \emph{pair} $P : A \to B$ is a tuple
$(A, B, T, \carry, \proj_P, d_P)$ where
$T : \Prog{A} \rightharpoonup \Prog{B}$ is the \emph{translator};
$\carry : \Beh{B} \rightharpoonup \Beh{A}$ is the
\emph{target-to-source interpreter}, which re-expresses a target behavior
as a source behavior; $\proj_P \subseteq F_A$ is the pair's
\emph{declared projection} --- exactly the source observables it promises
to preserve; and $d_P \in \{\mathsf{exact}, \mathsf{over}\}$ is its
\emph{declared direction}. An $\mathsf{over}$ (over-approximating) pair
additionally ships a pure \emph{witness embedding} $W$, mapping each
closing valuation $x$ of a source program $p$ to a closing valuation
$W(p,x)$ of $T(p)$; for an $\mathsf{exact}$ pair, $W$ is the identity.
$T$, $\carry$, and $W$ are pure (as in \Cref{def:interpreter});
$I_A$ and $I_B$ are owned by the languages and shared by every pair that
touches them. The pair's domain $\dom(P)$ is the set of $p$ with
$p \in \dom(I_A) \cap \dom(T)$, $T(p) \in \dom(I_B)$, and
$I_B(T(p)) \in \dom(\carry)$. Like the projection, the direction is a
declared, \emph{protected} part of the contract.
\end{definition}

\begin{definition}[Faithfulness: the directional square]
\label{def:faithful}
\label{def:lax}
$P$ is \emph{faithful at} $p \in \dom(P)$ when its square commutes at
$p$ \emph{along the witness embedding}: for every closing valuation $x$
of $p$,
\[
\proj_P\bigl(I_A(p(x))\bigr) \;=\;
\proj_P\bigl(\carry(I_B(\,T(p)(W(p,x))\,))\bigr),
\]
where a closed program has the single trivial valuation. For an
$\mathsf{exact}$ pair ($W = \mathrm{id}$) this is the familiar
commuting square $\proj_P(I_A(p)) = \proj_P(\carry(I_B(T(p))))$:
\[
\begin{tikzcd}[column sep=large]
p \arrow[r, "T"] \arrow[d, "I_A"'] & T(p) \arrow[d, "I_B"] \\
I_A(p) & \carry(I_B(T(p))) \arrow[l, "\eqproj{\proj_P}"']
\end{tikzcd}
\]
$P$ is faithful on $C \subseteq \dom(P)$ when it is faithful at every
$p \in C$.
\end{definition}

The direction is the \emph{reading} of that equality. The embedding
selects which target valuations the square checks: exactly those in its
range. When $W$ is the identity --- surjective onto the closing
valuations of $T(p)$ --- every target valuation is checked and the
target has no behaviors beyond the source's kept ones: the square is
two-sided, a bisimulation on the kept observables, and that is all
$d_P = \mathsf{exact}$ asserts. When $W$ is not surjective, the target
valuations outside its range are the pair's \emph{added} behaviors ---
their existence is the point, not a defect --- and the same checked
equality asserts only the simulation half,
$I_A(p) \sqsubseteq_{\proj_P} \carry(I_B(T(p)))$: every source behavior
has a target counterpart on the kept observables, and the target may
have more. Exactness is thus not a separate notion but the
identity-embedding special case of one square (\Cref{prop:exactid},
machine-checked). The check itself is direction-indifferent --- run
both legs, compare under $\proj_P$ --- so everything below applies to
both directions verbatim; only what agreement \emph{means} differs,
which is why direction is protected like the projection: an
$\mathsf{over}$ square passed off as $\mathsf{exact}$ would launder
added behaviors into a meaning-preservation claim.

Three remarks. First, everything in \Cref{def:faithful} is computable:
faithfulness at a given $p$ is a \emph{decidable, executable check} --- run
both sides, compare under $\proj_P$. We call this check the
\emph{square oracle}; in fault-tolerance terms it is an acceptance test
in the sense of recovery blocks \citep{randell1975recovery}, turning a
silently wrong --- in effect Byzantine --- translator into a
\emph{fail-stop} component \citep{schlichting1983failstop}. Second, when the check fails it fails at a
\emph{first} position: a least step index $i$ and a field
$f \in \proj_P$ on which the two sides differ. The oracle reports
$(i, f)$ --- a translator (or interpreter, or carry-back) defect, localized
to a step and a named observable. Third, the projection is part of the
contract, not a tuning knob: $\proj_P$ states what ``preserves meaning''
promises for this pair, and $F_A \setminus \proj_P$ --- the pair's
\emph{loss} --- states what it does not.

The carry-back $\carry$ is what makes a target-side result \emph{mean}
something at the source: a solver counterexample or a target trace is
re-expressed as a source behavior. It is pair-specific because the
correspondence it encodes is pair-specific; it is also why the square's
oracle is as expressive as the translator it checks.

\begin{genexample}[One square, run --- and failing]
The \texttt{JAL} probe of the RV64IMC inventory
(\Cref{def:coverage}) is a two-instruction program whose taken jump
leaves the code image:
\begin{center}
\footnotesize\ttfamily
\begin{tabular}{@{}lll@{}}
0x0: & jal x1, 8 & ; link x1 := 4, jump to 0x8 \\
0x4: & ecall     & ; (jumped over) \\
\end{tabular}
\end{center}
The left leg runs the shared RISC-V interpreter $I_A$; the right leg
translates to BTOR2, interprets, and carries the trace back, compared
under $\proj \supseteq \{\texttt{pc}, \texttt{x1}, \texttt{halted}\}$.
The square is exact ($W = \mathrm{id}$) --- as is every registered
pair's but the abstraction endo-pairs' (\S\ref{sec:lax}).
Post-step states, with the right leg as the
\texttt{riscv-btor2} translator emitted them at version 0.1:
\begin{center}
\footnotesize
\begin{tabular}{@{}clll@{}}
\toprule
step & field & $I_A(p)$ & $\carry(I_B(T(p)))$ \\
\midrule
0 & \texttt{pc}, \texttt{x1}, \texttt{halted} & 8, 4, F & 8, 4, F \\
1 & \texttt{pc}, \texttt{x1} & 8, 4 & 8, 4 \\
  & \texttt{halted} & \textbf{T} & \textbf{F} \\
\bottomrule
\end{tabular}
\end{center}
Address \texttt{0x8} matches no instruction: the reference semantics
halts on the fetch miss; the translator had modeled only the
instructions. The oracle reports $(i, f) = (1, \texttt{halted})$ and,
run hop by hop (\Cref{cor:localization}), indicts the
\texttt{riscv-btor2} hop --- incident I21 (\S\ref{sec:bugs}), caught
by the first conjoined coverage run in three independently written
lowerings the same morning, fixed by a versioned translator bump with
the taken-jump squares as permanent regressions. The all-pass rows of
\Cref{tab:capability} are made of exactly such checks.
\end{genexample}

\subsection{Pasting: when squares compose}
\label{sec:pasting}

Two pairs compose when the middle language is shared:
$P_1 = (A, B, T_1, \carry_1, \proj_1, d_1)$ and
$P_2 = (B, C, T_2, \carry_2, \proj_2, d_2)$ yield the candidate composite
\[
P_2 \circ P_1 \;=\; (A,\, C,\; T_2 \circ T_1,\; \carry_1 \circ \carry_2,\;
\proj,\; d_1 \wedge d_2)
\]
for a projection $\proj \subseteq \proj_1$ to be determined, with
direction the meet on the chain $\mathsf{exact} > \mathsf{over}$ --- a
composite is exact iff both hops are --- witness embedding the
composition $W_2 \circ W_1$ (an exact hop contributing the identity),
and $\dom(P_2 \circ P_1)$ the evident composite domain. Folklore says the
squares paste. They do not paste for free: $P_2$ guarantees agreement of
the two $B$-behaviors only \emph{up to} $\proj_2$, and $\carry_1$ consumes
whole $B$-behaviors. If $\carry_1$ reads a $B$-field that $P_2$ discards,
the outer rectangle can fail while both inner squares hold. The missing
condition is that $\carry_1$ must not look outside what $P_2$ preserves:

\begin{definition}[Support]
\label{def:support}
$\carry_1$ is \emph{$(\proj_2 \Rightarrow \proj)$-supported} when for all
$b, b' \in \dom(\carry_1)$:
$\proj_2(b) = \proj_2(b')$ implies
$\proj(\carry_1(b)) = \proj(\carry_1(b'))$.
Equivalently: $\proj \circ \carry_1$ descends to a well-defined map on
$\proj_2$-equivalence classes --- a congruence condition, matching the
mechanization exactly.
\end{definition}

\begin{theorem}[Pasting]
\label{thm:pasting}
Let $p \in \dom(P_2 \circ P_1)$, let $P_1$ be faithful at $p$, let $P_2$
be faithful at $T_1(p)$, let $\proj \subseteq \proj_1$, and let
$\carry_1$ be $(\proj_2 \Rightarrow \proj)$-supported. Then the
composite $P_2 \circ P_1$ is faithful at $p$ with respect to $\proj$,
along the composed embedding $W_2 \circ W_1$ and with direction
$d_1 \wedge d_2$; for the exact--exact instance, writing
$r = T_2(T_1(p))$,
\[
\proj(I_A(p)) \;=\; \proj\bigl(\carry_1(\carry_2(I_C(r)))\bigr).
\]
\end{theorem}

\begin{proof}
Write $q = T_1(p)$. Then $\proj(I_A(p)) = \proj(\carry_1(I_B(q)))$ since
$\proj \subseteq \proj_1$ and $P_1$ is faithful at $p$. Faithfulness of
$P_2$ at $q$ gives
$\proj_2(I_B(q)) = \proj_2(\carry_2(I_C(r)))$, so the
support condition applies to the two $B$-behaviors and yields
$\proj(\carry_1(I_B(q))) = \proj(\carry_1(\carry_2(I_C(r))))$.
Chain the equalities. For the directional form, run the same chain per
closing valuation, the legs closed at $x$, $W_1(p,x)$, and
$W_2(T_1(p), W_1(p,x))$; mechanized as \texttt{lax\_pasting} and,
telescoped over routes with the direction meet,
\texttt{DRoute.lax\_route\_pasting} and
\texttt{DRoute.direction\_exact\_iff} (\S\ref{sec:mech}).
(Details: \Cref{app:proofs}.)
\end{proof}

The support condition is not decorative. Take $F_B = \{g, h\}$ with
$\proj_2 = \{g\}$, and let $\carry_1$ copy field $h$ into a
$\proj_1$-kept source field: $P_2$ can be faithful --- it preserves
$g$ --- while $\carry_2(I_C(r))$ differs from $I_B(q)$ on $h$, which
$\proj_2$ permits. The outer rectangle then fails on the copied field
although both inner squares commute.

\paragraph{Composed keep and loss}
In the implementable special case --- $\carry_1$ \emph{fieldwise} and
step-aligned, each source field $f$ computed from a dependency set
$\mathit{deps}(f) \subseteq F_B$ of target fields --- the support
condition has a syntactic reading, and the largest $\proj$ satisfying
\Cref{thm:pasting} is
\[
\keep(P_2 \circ P_1) \;=\;
  \{\, f \in \proj_1 \mid \mathit{deps}(f) \subseteq \proj_2 \,\},
\]
a source field survives the route iff $P_1$ keeps it and it is computed
only from target fields $P_2$ keeps. Dually
$\lost(P_2 \circ P_1) = F_A \setminus \keep(P_2 \circ P_1)$: route loss is
the union of per-hop losses, each pulled back along the carry-backs. This
is the formal content of the platform rule that a route must \emph{declare
its cumulative loss}: the observables a destination can still speak about
are exactly those no hop discarded. (When $\carry_1$ changes step
granularity --- one source step spanning several target steps, as in
C-to-ISA --- the same statement holds with $\mathit{deps}$ taken over the
spanned window; appendix~A.2.)

\begin{corollary}[Localization]
\label{cor:localization}
Under the support condition, if the composite square fails at $p$ (both
sides defined), then $P_1$'s square fails at $p$ or $P_2$'s square fails
at $T_1(p)$. Inductively, along a route
$R = P_n \circ \cdots \circ P_1$ satisfying the pairwise support
conditions, a failing outer rectangle implies a failing inner square at
some hop $i$ --- and running the square oracle hop by hop finds the
least such $i$, then the least failing step and field. (The check is
direction-indifferent, so this holds along any directional route ---
divergences localize the same way whether the failing square's reading
was $\equiv$ or $\sqsubseteq$.)
\end{corollary}

\Cref{cor:localization} is the debugging story of the whole platform: a
divergence anywhere along a route is never a mystery about the route; it
is an indictment of one hop, one step, one observable. It is also what
disagreement between \emph{routes} will lean on in \S\ref{sec:branching}.

\paragraph{Routes, and the contract algebra}
The registry of languages and pairs forms a directed multigraph; a
\emph{route} $R : A \to Z$ is a path through it, with composed translator
$T_R$, composed carry-back $\carry_R$, composed projection $\keep(R)$,
composed direction and embedding as above, and
faithfulness given by iterating \Cref{thm:pasting}. Composition, told
once: a pair's composable declaration is its \emph{contract} --- the
assurance class of its guarantee (\Cref{def:class}) paired with its
direction, alongside the keep-set and the measured cost --- and a
route's contract is the \emph{componentwise meet}, the weakest hop on
every axis at once, with nothing stronger entailed
(\texttt{Contract.comp\_glb}, machine-checked; the semantic content
of each coordinate is \Cref{thm:pasting} with
\Cref{prop:weakestlink}, and \Cref{prop:lax}). Purity composes too,
and gives caching:

\begin{proposition}[Determinism and caching]
\label{prop:cache}
If every component function of every hop of $R$ is pure, then $T_R$ and
$\carry_R$ are pure, and a content-addressed cache keyed by the input
hash and the component versions returns, on a hit, exactly what
recomputation would return --- across the whole route. If some hop is
impure, re-running the route on the same input and
diffing bytes at every hop (\emph{twice-and-diff}) can detect it and
localize it to the first hop whose bytes differ --- a sound but
incomplete detector: two coinciding runs of an impure hop pass.
\end{proposition}

\subsection{The direction axis: abstraction as a pair}
\label{sec:lax}

Every square so far shown ran exact, and most registered pairs do. The
$\mathsf{over}$ half of \Cref{def:faithful} earns its place on cost:
an \emph{abstraction} maps a program to a smaller or cheaper model with
\emph{more} behaviors --- cut a state variable's update and let it
range freely, and every source behavior survives while new, spurious
ones appear. Such a translation is exactly what makes an expensive
universal question tractable, and in this calculus it is not a special
mechanism bolted onto a solver run but an ordinary pair whose declared
direction is $\mathsf{over}$: same oracle, same coverage conjunction
(\Cref{def:coverage}), same ratchet (\Cref{prop:ratchet}), same
negative controls, checked along its witness embedding. Two results
pin the axis down: exactness is a special case, and universal verdicts
transfer.

\begin{proposition}[Exactness is the identity embedding]
\label{prop:exactid}
Under the specialization obligation of \Cref{thm:universal}(iv) ---
translating the closed instance agrees with closing the one open
translation --- a pair with $W = \mathrm{id}$ is faithful at an open
$p$ in the sense of \Cref{def:faithful} iff its two-sided square
commutes at every closed instance $p(x)$. The exact calculus of
\S\ref{sec:pairs}--\S\ref{sec:pasting} is thus the
$W = \mathrm{id}$ fragment of the directional one --- literally: the
mechanization derives each side from the other
(\texttt{laxFaithful\_id\_iff\_faithful}, axiom-free;
\S\ref{sec:mech}).
\end{proposition}

\begin{proof}
With $W = \mathrm{id}$ the two legs close at the same valuation, and
the specialization obligation identifies the closed translation
$T(p(x))$ with the closed instance $T(p)(x)$ of the open artifact;
the quantification over valuations is then the quantification over
closed instances. \qed
\end{proof}

\begin{proposition}[Universal transfer across over-approximation]
\label{prop:lax}
Let $R : A \to Z$ have composed direction $\mathsf{over}$ and be
faithful (\Cref{def:faithful}) on a fragment containing $p$, with
$\varphi$ over $\keep(R)$. If no valuation of $T_R(p)$ exhibits
$\varphi$ within the declared bounds, then no valuation of $p$ does:
\emph{universal verdicts transfer across over-approximation}. An
existential verdict does not transfer; it remains subject to the
carried-back replay of \Cref{thm:existential}, whose failure on an
$\mathsf{over}$ route has a second, benign reading --- a \emph{spurious
counterexample}, a demand to refine the abstraction rather than an
indictment of a hop.
\end{proposition}

\begin{proof}
Contrapositive: a valuation $x$ with $\varphi$ observed in
$\sem{p(x)}{A}$ yields, by faithfulness along the composed embedding,
the valuation $W_R(p,x)$ of $T_R(p)$ exhibiting $\varphi$ on the kept
observables --- present within bounds, contradicting the premise.
(Mechanized: \texttt{lax\_universal\_transfer}, \S\ref{sec:mech}.)
\end{proof}

Together with \Cref{thm:existential,thm:universal} this closes the
CEGAR loop with durable artifacts: a transferred \textsc{unreachable}
answers at the source (\Cref{prop:lax}); a \textsc{reachable} either
replays (a true counterexample) or is spurious --- and the refinement
the failure demands is a \emph{new registered pair} with a smaller
havoc set, graded and reusable. The artifact instantiates this with
the endo-pair \texttt{btor2-havoc}, localization abstraction on the
bit-level hub, its lax square, negative controls, and refinement loop
in the repository's gate (a second $\mathsf{over}$ endo-pair, interval
abstraction, is registered as a brief); \S\ref{sec:eval-abstraction}
measures the loop, on authored systems and on an ingested HWMCC
slice.

\paragraph{Advising the dial}
The abstraction's parameters stay the player's, but they need not be
guessed. The artifact ships an advisor over the bit-level hub: the
\emph{cone of influence} of a question --- rooted in its bad
conditions \emph{and every environment constraint} (a state gating
run validity is never free), closed backwards through the
state-evolution supports --- splits the state space into a
\emph{free} havoc set, whose abstraction provably cannot move the
question's signal, and a refinement ladder ordered
farthest-from-the-question first; observed per-state bounds seed
candidate intervals that the lax square then corroborates or refutes.
The free-set claim is executable and negative-controlled in the
artifact's tests; mechanizing it is a natural next entry for
\S\ref{sec:mech}'s audit.

\paragraph{The dual, for symmetry}
The axis has an evident third value the platform deliberately does not
yet inhabit: an \emph{under}-approximating square --- the embedding
running the other way, the target with \emph{fewer} behaviors ---
where existential verdicts would transfer and universal ones would
not, the mirror image of \Cref{prop:lax}. Nothing in this paper
depends on it; we note it because the directional reading makes the
symmetry visible, and its verdict-transfer discipline would fall out
of the same monotonicity.

The mechanization of \S\ref{sec:mech} covers the directional calculus
in full: the directional square, pasting along the composed embedding
(binary and telescoped) with the direction meet, and both propositions
above are machine-checked (\texttt{Calculus/Lax.lean}) with the same
axiom footprint as the exact fragment.

\subsection{Contracts: grades, classes, and the meet}
\label{sec:fidelity}
\label{sec:grades}

A pair declares a \emph{fidelity grade} from
$G = \{$\ttrust, \trepr, \tchk, \tpred, \tprov$\}$ --- the
\emph{evidence species} behind its guarantee: a spec audit that
re-derives the bytes ($\tpred$), an independently re-checked
certificate ($\tprov$), this run's square-oracle verdict ($\tchk$), a
digest pin ($\trepr$), or nothing ($\ttrust$). Grades differ in kind
($\tpred$ and $\tprov$ are incomparable: foreseeable vs.\
re-checkable); what composition needs is the logical \emph{form} of
the guarantee, and that is totally ordered:

\begin{definition}[Assurance class]
\label{def:class}
The \emph{assurance classes} are the chain
$\mathbb{A} : \mathsf{none} < \mathsf{replay} < \mathsf{perrun} <
\mathsf{universal}$, and
$\mathit{class} : G \to \mathbb{A}$ maps
$\ttrust \mapsto \mathsf{none}$,
$\trepr \mapsto \mathsf{replay}$,
$\tchk \mapsto \mathsf{perrun}$, and
$\tpred, \tprov \mapsto \mathsf{universal}$.
The classes denote guarantee predicates for a pair $P$ with covered
fragment $C_P$:
$\mathsf{universal}$ --- $P$ is faithful at every $p \in C_P$;
$\mathsf{perrun}$ --- $P$ is faithful at every $p$ on which the square
oracle has run and passed (in particular, at the program of the present
run);
$\mathsf{replay}$ --- $T$ is pure, nothing about meaning;
$\mathsf{none}$ --- nothing.
\end{definition}

Only classes compose: a route with one $\tpred$ hop and one $\tprov$
hop is $\mathsf{universal}$ end to end, as it should be.

Two honesty notes on the mapping. First, grade \emph{declarations} are
themselves trusted inputs: only $\tchk$ is mechanically enforced by the
framework (the oracle runs); $\mathit{class}(\tpred) = \mathsf{universal}$
additionally presumes that the audited specification maps rules to
\emph{semantics} --- a translator faithful to a wrong rule table is
$\tpred$ and universally unfaithful, and our own evaluation exhibits a
$\tpred$ edge that was wrong on covered inputs until a coverage probe
caught it (\S\ref{sec:bugs}). Declarations therefore sit in the
trusted-base ledger like any other component. Second, $\tprov$ as
instantiated in \S\ref{sec:eval-proved} certifies a \emph{per-question}
verdict --- universal over that question's inputs, per-run over
questions --- not the pair's square for all programs.

\label{sec:composition}

\begin{proposition}[Weakest link]
\label{prop:weakestlink}
Let route $R = P_n \circ \cdots \circ P_1$ satisfy the support conditions
of \Cref{thm:pasting}. If every hop's guarantee holds, then the composite
guarantee of class $\min_i \mathit{class}(P_i)$ holds for $R$ (with
$C_R$ the composite fragment), and no higher class is entailed by the
hop guarantees alone.
\end{proposition}

The proof is immediate from \Cref{def:class} and \Cref{thm:pasting}:
universal statements conjoin into universal statements over the composite
fragment; a per-run statement at hop $i$ caps the conjunction at the
programs the oracle actually saw; a $\mathsf{replay}$ hop caps it at
purity. Optimality is by the evident counterexamples. Statically, then,
\emph{a route is as faithful as its weakest hop} --- and this is one
instance of the paper's single composition statement: a pair's
composable declaration, its \emph{contract} (assurance class $\times$
direction), composes along a route by the componentwise meet, the
weakest hop on every axis at once (mechanized:
\texttt{Contract.comp\_glb}, \S\ref{sec:mech}; the class coordinate is
this proposition, the direction coordinate is \Cref{prop:lax}). The
static story is still not the whole story, because every pair carries
its own oracle:

\begin{theorem}[Per-run re-establishment]
\label{thm:reestablish}
Fix a run of route $R$ on program $p$, with intermediate programs
$q_0 = p$, $q_i = T_i(q_{i-1})$. If on this run the square oracle runs
and passes at every hop $i$ (on $q_{i-1}$), and the support conditions
hold, then $R$ is faithful at $p$ --- regardless of the hops' static
grades. In particular a $\trepr$ or even $\ttrust$ hop that is
dynamically validated at $q_{i-1}$ contributes, for this run, exactly
what a $\tchk$ hop contributes.
\end{theorem}

\begin{proof}
Each passed oracle check \emph{is} faithfulness at $q_{i-1}$
(\Cref{def:faithful} is decidable and the oracle decides it); apply
\Cref{thm:pasting} inductively.
\end{proof}

\Cref{thm:reestablish} applies exactly to hops whose squares run: the
pinned C compiler at the spine's head has no square, so the inline
oracles lift the route \emph{from the compiled binary onward}, the
compiler contributing $\trepr$ plus a corpus-level differential
against an independent C verifier (\Cref{sec:instantiation}).
Re-establishment is per-run evidence, never a static promotion, and
never coverage for a hop with no oracle.

\subsection{Branches, diversity, and the trusted base}
\label{sec:branching}

The registry graph is not a line: one source may reach one destination by
several routes. Let $R_1, R_2 : A \to Z$ with composed keep-sets
$K_1, K_2$ and $K = K_1 \cap K_2$, sharing the endpoint interpreters
(language-owned) but no translator or carry-back.

\begin{lemma}[Disagreement localizes]
\label{lem:disagree}
If $R_1$ and $R_2$ disagree at $p$ under $K$, at least one route is
unfaithful at $p$, and \Cref{cor:localization} pins the failure to a hop,
a step, and a field of that route.
\end{lemma}

\begin{lemma}[Agreement corroborates]
\label{lem:agree}
If $R_1$ and $R_2$ agree at $p$ under $K$, then either both are faithful
at $p$ (under $K$), or both are unfaithful with \emph{identical projected
error}: defects in independently built components producing the same
wrong $K$-behavior at $p$.
\end{lemma}

Both are immediate: two faithful routes each equal $K(I_A(p))$, hence
each other. What \Cref{lem:agree} buys depends on an assumption we state
rather than hide:

\begin{assumption}[Diversity]
\label{asm:diversity}
Corroborating routes share no translator or carry-back \emph{on the
segment where they diverge} (up to the language at which they
reconverge), and the diverse translators are derived from
\emph{independent semantic artifacts} --- e.g.\ RISC-V reaches BTOR2
once via a translator written from the prose ISA specification and once
via the Sail formal model; likewise AArch64 from the Arm manual vs.\ the
Arm Sail model.
\end{assumption}

The assumption's premise is, moreover, a \emph{declared,
protected registry fact} rather than prose alone: each pair names the
semantic artifact its translator derives from, and a mechanical
judgment (the trust advisor of \S\ref{sec:books}) scores every branch
by the diverse segments' declared artifacts, the shared suffix
removed --- a shared artifact is never independent, an undeclared one
is unknown, never silently independent --- and states saturation
plainly: pairs can be generated without bound, independent artifacts
cannot, and corroboration stops growing at their supply.

The scoping matters, and independence must be argued, not assumed. Below
the reconvergence, routes may share hops --- in the platform both
RISC-V$\to$SMT-LIB routes share the BTOR2$\to$SMT-LIB bridge --- and
shared emission libraries are shared failure surface, so agreement
corroborates the diverse prefix and nothing beneath it. The evaluation
therefore also runs a \emph{disjoint-decision} variant of every branch
question (\S\ref{sec:eval-branch}, \Cref{fig:branch}): the direct route's BTOR2 system
decided natively by btormc, against the Sail-derived route through the
bridge and Z3, so the diverse segment extends through the decision
procedure and the shared bridge drops out of the corroborating pair
entirely; what remains shared is the emission library (the site of the
cross-route incident below) and the language-owned endpoints. Correlated
failures between nominally independent implementations are, moreover,
the empirical norm \citep{knight1986independence}; our own history
exhibits both a cross-route common-mode failure through a shared emitter
and a shared-misreading blind spot (\S\ref{sec:bugs}), and translators
written by LLM agents plausibly share model-family failure modes.

Under \Cref{asm:diversity}, the residual failure mode of an agreeing
branch is a \emph{common-mode failure}: independently produced encodings
of the same semantics, wrong in the same observable way on the same
program. We do not assign this a probability; we account for it
structurally, in the trusted base ledger below. This is how the platform
manufactures assurance out of hops that are individually only
$\tchk$ --- dual modular redundancy whose comparator, unlike a voter,
localizes: agreement of independently derived routes is evidence that
neither route's grade alone provides.

\label{sec:tcb}

Define the \emph{trusted computing base} $\tcb(v)$ of a verdict $v$ as
the set of components whose correctness $v$'s soundness still assumes
after all checks that ran. Each cross-check layer strictly shrinks it
(\Cref{tab:ledger}); the last two rows are not interchangeable, and the
difference is the central asymmetry of the whole design.

\begin{table}
\caption{The trusted-base ledger: what each layer of cross-checking
removes from the set of components a verdict still assumes correct.}
\label{tab:ledger}
\footnotesize
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.36\linewidth}%
>{\raggedright\arraybackslash}p{0.25\linewidth}%
>{\raggedright\arraybackslash}p{0.30\linewidth}@{}}
\toprule
\textbf{Layer that ran} & \textbf{Removed from \tcb} &
\textbf{Still in \tcb} \\
\midrule
bare route + solver & --- &
all $T_i$, $\carry_i$, all interpreters, solver \\
+ square oracle at every hop (\Cref{thm:reestablish}) &
every $T_i$ &
the $\carry_i$, the interpreters, solver \\
+ branch agreement (\Cref{lem:agree}, \Cref{asm:diversity}) &
the diverse prefix's $\carry_i$ (cross-corroborated) &
shared endpoint interpreters, shared suffix hops and emission
libraries, solver; common-mode residue \\
+ independent solver corroboration & the single solver &
the agreeing engines' common failure modes \\
+ certificate check ($\tprov$; e.g.\ DRAT/LRAT) & the solver entirely &
the certificate checker (formally verified, in the \texttt{cake\_lpr}
instantiation), plus the bit-blaster's BV$\to$CNF step \\
+ source-level witness replay (\S\ref{sec:endtoend}) &
\emph{everything above} &
$I_A$ adequacy (\Cref{asm:adequacy}) + the replay harness \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The end-to-end asymmetry}
\label{sec:endtoend}

Finally, what does a user hold at the end? Let $R : A \to Z$ be a route
into a \emph{reasoning language} $Z$ (one a solver consumes: a
transition-system logic, an SMT theory). The user asks a question about
an \emph{open} source program $p(x)$ --- free inputs $x$ --- and a
condition $\varphi$ over $\keep(R)$-observables: is $\varphi$ reachable?
The solver decides at $Z$; its answer comes back in one of two logical
shapes, and they differ fundamentally in what must be trusted.

\begin{theorem}[Existential answers are self-certifying]
\label{thm:existential}
Suppose the solver returns \textsc{reachable} with witness $w$, the
composed carry-back $\carry_R$ re-expresses $w$ as an input valuation
$x_0$ (and claimed trace), and replaying $I_A(p(x_0))$ exhibits
$\varphi$. Then $\varphi$ is truly reachable in $\sem{p}{A}$ --- with
$\tcb = \{$\Cref{asm:adequacy} for $I_A\}$ plus the small fixed replay
harness the judgment runs through (the projection, the
$\varphi$-evaluator, and the substitution closing $p$), and nothing
else. Every
translator, carry-back, hub interpreter, and solver on $R$ is a
\emph{discovery} device only: a defect in any of them can cause a missed
or non-replaying witness, never a false \textsc{reachable}.
\end{theorem}

\begin{proof}
The replay is a run of $I_A$ itself on a concrete closed program
$p(x_0)$; by \Cref{asm:adequacy} its behavior is $\sem{p(x_0)}{A}$, and
$\varphi$ was observed on it. No other component's output enters the
final judgment.
\end{proof}

\begin{theorem}[Universal answers need the machinery]
\label{thm:universal}
Suppose the solver returns \textsc{unreachable} (within declared
unrolling/step bounds $k$). If (i) every hop of $R$ is faithful at every
instance $p(x)$ --- the $\mathsf{universal}$-class guarantee on the
fragment containing $p$'s constructs, and exactly the hypothesis the
mechanization takes; (ii) the support conditions of
\Cref{thm:pasting} hold and $\varphi$ mentions only $\keep(R)$;
(iii) the verdict is certified by an independently checked certificate
or corroborated across independent engines; and (iv) the symbolic
artifact the solver consumes denotes the family of closed translations
the squares check --- translation commutes with input specialization, an
obligation each reasoning pair discharges by construction of its
encoding and, where the encoding carries free inputs, by sampled
commutation tests (\S\ref{sec:mech}), and under which the mechanization
derives every
per-instance translation from the one open translation the platform
actually performs (\texttt{universal\_from\_open\_artifact},
\S\ref{sec:mech}) --- then no input drives
$\sem{p}{A}$ to $\varphi$ within $k$ steps, with $\tcb$ as in the
corresponding row of the ledger (\S\ref{sec:tcb}).
\end{theorem}

One caution on hypothesis (i): per-run oracle evidence combined with
branch agreement (\Cref{lem:agree}, \Cref{asm:diversity}) \emph{raises
confidence} in it but does not entail it --- oracle passes cover the
finitely many closures actually run, and agreement bounds the residual
failure to common-mode rather than excluding it. Universal verdicts are
precisely where the $\mathsf{universal}$ tiers earn their keep; evidence
below that tier is corroboration, not entailment.

Direction slots into the same asymmetry monotonically: along an
$\mathsf{over}$ route a certified \textsc{unreachable} transfers to the
source (\Cref{prop:lax}), while a \textsc{reachable} already rested on
the carried-back replay --- so the directional calculus adds no new
trust obligation, only a second, benign reading of a replay failure:
a spurious counterexample, the refinement demand of \S\ref{sec:lax}.

The asymmetry between \Cref{thm:existential,thm:universal} is, we argue,
the right lens on the entire design: \emph{for questions whose answers
carry witnesses, trust is nearly free} --- carry the witness back and
replay it at the source; the platform's graded tiers, branches, and
certificates exist for the other half, the universal answers where no
witness can vouch for the verdict. This is the end-to-end argument
\citep{saltzer1984endtoend} transplanted to translation: the check that
settles a witness-carrying answer belongs at the endpoint --- the
source interpreter --- and what the intermediate hops provide is, for
that half, an optimization; only for universal answers, where no
endpoint check can exist, does hop fidelity become the guarantee
itself. A system that makes the carry-back
$\carry$ a first-class, per-pair component is a system in which the cheap
half of trust is always cashed in; a system that grades and stacks
cross-checks is one in which the expensive half is bought consciously,
in labeled increments, with the bill itemized in the
ledger~of~\S\ref{sec:tcb}.

\subsection{Mechanization}
\label{sec:mech}

The compositional core of \Cref{sec:calculus,sec:fidelity} is
mechanized in Lean~4 (core library only, no mathlib; $\sim$820 lines,
in the artifact): the pasting theorem with the support condition,
localization, the assurance-class chain and weakest link (sound half),
per-run re-establishment, both branch lemmas, the ratchet (verdict
preservation and coverage monotonicity under extensions), both
end-to-end theorems, the \emph{specialization obligation} of
\Cref{thm:universal}(iv) --- with the theorem that it discharges the
per-instance-translation hypothesis from the single open translation
the platform performs --- and the \emph{route telescope} --- routes of
arbitrary length as a dependently-typed chain of pairs over
interpreter-bundled languages, with the paper's ``pairwise support
conditions'' as a recursive coherence predicate, and the $n$-ary
pasting and localization theorems proved by induction through the
binary ones (the canonical head-projection choice at each junction is
shown lossless by a reprojection lemma). Three modeling
choices mirror the paper exactly. Components are Lean functions, so
\emph{purity is by construction} and \Cref{prop:cache} is
definitional; partiality is \texttt{Option}, with \unsupported{} as
\texttt{none}; and projections are generalized from field subsets to
arbitrary observable-compression maps, with ``$\proj \subseteq
\proj_1$'' becoming \emph{factoring}. Because every theorem takes each
component-output witness as an explicit hypothesis, a statement's
trusted base is literally its hypothesis list ---
\Cref{thm:existential} in Lean assumes source-interpreter adequacy and
nothing else, and its proof term is three tokens:
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize,
  literate={∀}{{$\forall$}}1 {→}{{$\to$}}1 {₀}{{$_0$}}1
           {φ}{{$\varphi$}}1 {π}{{$\pi$}}1 {∃}{{$\exists$}}1
           {∧}{{$\land$}}1 {⟨}{{$\langle$}}1 {⟩}{{$\rangle$}}1]
theorem existential_self_certifying
  (adequacy : ∀ p b, IA p = some b → sem p = some b)
  (hreplay : IA p₀ = some b) (hφ : φ (projB π b)) :
  ∃ b', sem p₀ = some b' ∧ φ (projB π b') :=
  ⟨b, adequacy _ _ hreplay, hφ⟩
\end{lstlisting}
\Cref{tab:mech} maps the paper's results to their Lean theorems with
each proof's exact axiom footprint, re-checked at every build; the one
classical proof is \Cref{cor:localization} (an unfaithful route names
no witness by itself). The table also shows where mechanization stops,
and we state it plainly: \Cref{thm:universal}'s clauses (iii) and (iv)
enter the Lean statement as \emph{hypotheses}. Both now carry
\emph{tested surrogates} rather than by-construction assertions. The
specialization obligation is discharged by sampled commutation checks
per reasoning pair with free inputs (in the suite): the open program is
translated once, and for sampled valuations the emitted artifact is
byte-identical while the specialized instance's square passes --- the
shape \texttt{universal\_from\_open\_artifact} consumes. The
solver-artifact-to-target-semantics correspondence behind (iii) is
corroborated per verdict: every bounded-unreachable route verdict in
the branch, benchmark, and certified experiments is additionally
replayed through the strict
target interpreter for the full bound --- sampled inputs where the
system has any --- and no \texttt{bad} may fire
(\S\ref{sec:eval-branch}--\S\ref{sec:eval-proved}). Sampling
corroborates, it does not entail --- which is the grade discipline
applied to the calculus's own hypotheses.

\begin{table}
\caption{The mechanization at a glance: paper result, Lean theorem,
and the proof's axiom footprint as printed by the build-time audit
(\texttt{Audit.lean}); ``---'' is axiom-free.}
\label{tab:mech}
\footnotesize
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.27\linewidth}%
>{\raggedright\arraybackslash}p{0.42\linewidth}%
>{\raggedright\arraybackslash}p{0.22\linewidth}@{}}
\toprule
\textbf{Result} & \textbf{Lean theorem} & \textbf{Axioms} \\
\midrule
\Cref{thm:pasting} (pasting) & \texttt{pasting}, \texttt{pasting}$_3$
  & \texttt{propext} \\
\Cref{cor:localization} & \texttt{localization}
  & \texttt{propext}, \texttt{Classical.choice}, \texttt{Quot.sound} \\
\Cref{prop:weakestlink} (sound half) & \texttt{weakest\_link\_universal}
  & \texttt{propext} \\
\Cref{thm:reestablish} & \texttt{reestablishment} & \texttt{propext} \\
\Cref{lem:disagree,lem:agree} & \texttt{disagreement\_localizes},
  \texttt{agreement\_corroborates} & --- \\
\Cref{prop:ratchet} & \texttt{ratchet\_preserves\_faithful},
  \texttt{ratchet\_coverage\_mono} & --- \\
\Cref{thm:existential} & \texttt{existential\_self\_certifying} & --- \\
\Cref{thm:universal} & \texttt{universal\_needs\_machinery}
  & \texttt{propext} \\
\Cref{thm:universal}(iv) & \texttt{universal\_from\_}\newline
  \texttt{open\_artifact} & \texttt{propext} \\
route telescope ($n$-ary) & \texttt{route\_pasting},
  \texttt{route\_localization} & \texttt{propext}, \texttt{Quot.sound}
  (+ \texttt{Classical.choice} for localization) \\
directional pasting, direction meet (\Cref{thm:pasting}) &
  \texttt{lax\_pasting}; telescoped:
  \texttt{DRoute.lax\_route\_pasting},
  \texttt{DRoute.direction\_exact\_iff} & \texttt{propext}
  (+ \texttt{Quot.sound} telescoped) \\
\Cref{prop:exactid} (exactness $=$ identity embedding) &
  \texttt{laxFaithful\_id\_iff\_faithful} & --- \\
\Cref{prop:lax} (universal transfer) &
  \texttt{lax\_universal\_transfer} & \texttt{propext} \\
contract algebra (composition = componentwise meet) &
  \texttt{Contract.comp\_glb}, \texttt{Direction.comp\_eq\_min}
  & \texttt{propext}, \texttt{Quot.sound} \\
\bottomrule
\end{tabular}
\end{table} Deliberately not mechanized: the fieldwise loss
computation and retiming case of appendix~A.2\ (syntactic
side
conditions over a concrete carry-back representation), the optimality
halves of \Cref{prop:weakestlink} (meta-statements over models), and
the grade set $G$ itself --- evidence provenance is not a mathematical
object, which is the point of separating $G$ from $\mathbb{A}$. The
telescope theorems' audit adds only \texttt{Quot.sound} (from
structural-recursion equations); localization remains the sole
classical proof.
