\section{Limitations and conclusion}
\label{sec:conclusion}

\paragraph{Limitations}
This is a dated snapshot of a system designed to ratchet, and its
numbers say so: coverage is deep on the spine (all of RV64IMC surviving
composed to SMT-LIB; the in-scope eBPF set complete) and deliberately
partial elsewhere (EVM at 86 of 144 opcodes; Wasm without
\texttt{loop}/\texttt{br}; Python an integer subset with bounded
unrolling). Universal answers are bounded claims wherever BMC-style
unrolling caps apply. The \tprov\ tier's certified row is inhabited with a \emph{formally
verified} checker at the anchor (\texttt{cake\_lpr},
\S\ref{sec:eval-proved}); the residual \tcb\ is the bit-blaster's
BV$\to$CNF step, and a certified claim is per-question and bounded,
never a proof of a pair.
Interpreter adequacy remains an assumption discharged empirically, and
the diversity assumption behind branch corroboration is structural, not
probabilistic. The compositional core of the calculus --- including the $n$-ary route
telescope --- is mechanized in Lean~4 with a clean axiom audit
(\S\ref{sec:mech}); the fieldwise loss computation and the retiming
case remain paper-proved.

\paragraph{Conclusion}
We proposed treating translations the way distributed systems treat
processes: as components that fail, whose failures are contained by
architecture --- declared projections, decidable squares, graded
evidence, independent routes, carried-back witnesses --- rather than
prevented by universal proof. The calculus makes the composition of
partial trust exact; the platform shows the discipline is buildable, by
agents, at useful coverage; and the asymmetry theorems say where proof
effort actually buys assurance. The snapshot is the baseline; the
ratchet is the roadmap.
