\section{Evaluation}
\ifarxiv
\noindent The ten subsections below answer four questions. \emph{Does
the graph answer correctly?} --- the capability snapshot, composed
coverage and branch agreement, the compliance slice, and the end-to-end
case studies (\S\ref{sec:eval-bench}--\S\ref{sec:eval-cases}).
\emph{Does the evidence mean what it claims?} --- the certified tier
and the escape-rate experiments for the gate itself
(\S\ref{sec:eval-proved}, \S\ref{sec:eval-escape}). \emph{Can
untrusted authors build it?} --- the defects the architecture caught
(\S\ref{sec:bugs}). \emph{Can an LLM play it?} --- the player
experiment (\S\ref{sec:eval-player}); performance and scale close
(\S\ref{sec:eval-perf}). Post-snapshot exhibits, explicitly not
benchmark claims, end the section (\S\ref{sec:eval-exhibits}).
\fi
\label{sec:evaluation}

%% Numbers in this section are generated into results/tables/ by
%% results/harvest.py; see paper/README.md. Env: results/data/env.json.

We evaluate the snapshot along the axes the calculus makes measurable:
per-pair coverage (\S\ref{sec:eval-capability}), composed route
coverage and branch agreement --- including a disjoint-decision block
--- (\S\ref{sec:eval-branch}), a
compliance-derived reachability benchmark decided along both RISC-V
routes (\S\ref{sec:eval-bench}), end-to-end case
studies exercising \Cref{thm:existential} (\S\ref{sec:eval-cases}),
certified-unreachability runs inhabiting the \tprov\ tier
(\S\ref{sec:eval-proved}), the defects the architecture caught
(\S\ref{sec:bugs}), a measured escape-rate estimate for the gate
(\S\ref{sec:eval-escape}), a first controlled experiment with LLM
players (\S\ref{sec:eval-player}), and performance, determinism, and
scale (\S\ref{sec:eval-perf}). Every number in this section's tables is
regenerated by one script from
the live registry and real runs at the snapshot commit, with a few
curated exceptions marked at source --- the defect catalog behind
\Cref{tab:bugs} is hand-mined from history, and \Cref{tab:player}
formats the recorded outputs of a manual two-arm protocol; the escape
experiment's earlier-round numbers (\S\ref{sec:eval-escape}) likewise
come from its incident record. (Host:
macOS/arm64, CPython 3.12, Z3 4.13, btormc 3.2, cadical; drat-trim
pinned at source commit \texttt{2e3b2dc} and cake\_lpr at
\texttt{a4323b2}, the engine inventory recorded in the regenerated
\texttt{env.json}; the platform's own suite runs 1215 tests there, all
green, with 3 host-dependent skips.) The evaluation remains deliberately
\emph{fully specified} rather than large: every question --- authored
or derived --- is checked end to end with its ground truth stated, and
the largest sweep (the 78-question compliance-derived benchmark of
\S\ref{sec:eval-bench}) is still small next to an SV-COMP-scale
campaign, which remains future work.

\subsection{Capability snapshot}
\label{sec:eval-capability}

\input{results/tables/capability}

\Cref{tab:capability} reports per-construct coverage measured at
harvest time --- not transcribed from documentation --- in both of
\Cref{def:coverage}'s readings: \emph{accepted} (the probe translates
without a typed $\unsupported$ abort) and \emph{conjoined} (accepted
\emph{and} the pair's square oracle passes on the probe --- the
definition's actual conjunction, run probe by probe). For every pair
with a decidable square the two columns now coincide; they did not
when the conjunction was first measured, and what it surfaced on its
first run is instructive enough to report
(\S\ref{sec:bugs}, incidents I20--I22): the Sail-route front dropped
the program's \emph{initial memory}, so every load-family probe
diverged at step 0 (accepted, silently wrong --- precisely the failure
mode acceptance counting cannot see), and three of the BTOR2 lowerings
modeled a program counter that leaves the code as \emph{stuck} where
every reference interpreter halts, so every taken-jump probe whose
target exits the image diverged on \texttt{halted} (the worked example
of \S\ref{sec:pairs} shows this square, and its oracle output, verbatim). It also audited its
own instrument: one probe (\texttt{SD}) stored over its own
\texttt{ECALL} terminator, so the reference interpreter could not run
it and no square had ever executed on that construct --- acceptance
counted it covered anyway. Measuring the
definition's conjunction found all three in its first run; the fixes
are ordinary versioned translator bumps and one probe repair. Pairs without a decidable square
--- the \texttt{predicted}-grade hops into the SMT-LIB hub --- discharge
the faithfulness
conjunct per run at question time instead (\S\ref{sec:composition}) ---
what their grade leaves to per-run checking; the table marks them
\emph{per-run} (the fourth \texttt{predicted} pair,
SMILES$\to$formula, has executable ends and conjoins directly). Denominators are language-owned inventories
(\Cref{def:coverage}'s yardstick): both RISC-V-headed pairs measure
against the same 96-construct RV64IMC inventory, the two AArch64 pairs
against one declared A64 slice whose out-of-scope constructs count in
the denominator, and the Wasm and EVM rows against their in-scope
$\cup$ enumerated-out-of-scope unions (54/75, 86/144) --- a gap shows
as a typed $\unsupported$ entry, never by shrinking the total. Depth
remains uneven by design and visibly so: the spine is deep (all 96
RV64IMC constructs conjoined; eBPF complete at 126; the
BTOR2$\to$SMT-LIB bridge total on its 56-item
operator/sort/directive inventory), Python stands at 11 of 27 subset
constructs, and every gap is a \emph{typed} $\unsupported$ the harness
can enumerate (the ``gaps'' column). The declared A64 slice is still
narrow against the full base ISA --- the honest reading of 27/33 is
``the in-scope family conjoins, and the slice is small.'' The C head
carries no construct inventory: its translator is the pinned compiler,
graded $\trepr$ (\S\ref{sec:composition}).

\subsection{Composed coverage and branch agreement}
\label{sec:eval-branch}

\input{results/tables/branch}

\begin{figure}
\centering
\begin{tikzpicture}[x=1cm, y=1cm]
\node[art]  (p)    at (3.4, 0)     {RISC-V program $p$ + question};
\node[lang] (b1)   at (1.4, -1.0)  {btor2};
\node[lang] (sail) at (5.4, -1.0)  {sail};
\node[lang] (b2)   at (5.4, -2.0)  {btor2};
\node[hub]  (smt)  at (5.4, -3.0)  {smtlib};
\node[tool] (mc)   at (1.4, -4.0)  {btormc};
\node[tool] (z3)   at (5.4, -4.0)  {z3};
\node[art]  (v1)   at (1.4, -5.0)  {verdict};
\node[art]  (v2)   at (5.4, -5.0)  {verdict};
\draw[chk]  (p) -- node[elbl, left=2pt]  {riscv-btor2} (b1);
\draw[chk]  (p) -- node[elbl, right=2pt] {riscv-sail}  (sail);
\draw[chk]  (sail) -- node[elbl, right]  {sail-btor2}  (b2);
\draw[pred] (b2) -- node[elbl, left]     {btor2-smtlib} (smt);
\draw[chk]  (b1) -- node[elbl, right=2pt, align=left, font=\tiny\itshape]
  {decide natively\\(no bridge, no\\SMT-LIB, no Z3)} (mc);
\draw[chk]  (smt) -- (z3);
\draw[chk]  (mc) -- (v1);
\draw[chk]  (z3) -- (v2);
\draw[<->, >=stealth] (v1) -- node[elbl, above] {must agree} (v2);
\end{tikzpicture}
\caption{What the bottom block of \Cref{tab:branch} measures: the same
question decided along two stacks sharing no \emph{decision} component
after the head (the named residue: emission library, endpoints) ---
the ISA-prose-derived lowering decided \emph{natively} by btormc,
against the Sail-model-derived lowering through the
\texttt{predicted}-grade bridge and Z3. Agreement corroborates the
whole diverse segment (\Cref{asm:diversity}); the stated residual
shared surface is the BTOR2 emission library and the language-owned
endpoint interpreters. The solver-level rows at the top of
\Cref{tab:branch} instead decide \emph{both} sides through the bridge
and Z3, sharing everything below the reconvergence.}
\label{fig:branch}
\end{figure}

\Cref{tab:branch} (top) composes conjoined coverage along whole routes:
a construct counts iff its probe survives \emph{every} hop and every
hop with a decidable square passes it on that hop's input. Nothing a
head accepts is eaten at the hubs: every accepted probe of every route
survives composed to its route's target, squares included. The denominators are
the source language's inventory, so the two RISC-V routes are measured
on the \emph{same} 96-construct yardstick and both compose 96/96 ---
the spine composes losslessly, and the earlier version of this table,
whose two RISC-V rows quoted different pair-chosen denominators
(96 vs.\ 95), is exactly the yardstick-shopping \Cref{def:coverage}
forbids and this measurement now closes. The AArch64 rows also carry a ratchet
anecdote the machinery makes precise: one revision earlier the
Sail-mediated route composed to $0/33$, because the Sail$\to$BTOR2
translator lowered only the RV64IMC arm --- and the harness localized
every miss to the \texttt{sail-btor2} hop by name. Widening that one
hop (an additive translator bump, RISC-V arm byte-identical per
\Cref{prop:ratchet}) closed the gap without re-earning any other
evidence. A gap the system \emph{reports and localizes} is the designed
behavior; a gap papered over would be the defect.

The bottom of \Cref{tab:branch} is the fidelity payoff of
\Cref{lem:agree}: the same reachability questions decided along
independently derived routes. Twelve solver-level questions ---
constant dataflow, summation and countdown loops under unrolling,
store/load through memory, two questions about a \emph{gcc-compiled C
program} (can \texttt{a0} be 47 at halt? can it be 99?), and four
AArch64 questions decided along the Arm-manual-derived and the
Arm-Sail-model-derived route --- were each decided along both routes of
their branch; all twelve pairs of verdicts agree, including the
universal (\textsc{unreachable}) halves where agreement carries real
information. The C rows exercise the full re-establishment story: an
opaque $\trepr$ compiler heads both routes, and its output is validated
downstream twice, independently. The AArch64 trace-level rows
cross-check the same branch below the solver, directly comparing the
two carried-back behaviors under $\proj$. The table's bottom block
addresses \Cref{asm:diversity}'s scoping head-on: the twelve
solver-level rows share the BTOR2$\to$SMT-LIB bridge below the
reconvergence, so every one of the twelve is re-decided with fully
disjoint \emph{decision} stacks after the head (\Cref{fig:branch}) --- the direct
route's BTOR2 system
decided \emph{natively} by btormc (no bridge, no SMT-LIB, no Z3)
against the Sail-model route through the bridge and Z3. All twelve
agree. Every unreachable verdict among them is additionally
corroborated below the solvers: the strict BTOR2 interpreter replays
the direct route's system for the full bound and no \texttt{bad} fires
--- the tested surrogate for \Cref{thm:universal}'s
artifact-to-semantics hypothesis (recorded per question, as is the
same replay for every bounded-unreachable verdict of
\Cref{tab:bench}). The diverse segment now spans the lowering derivation
(ISA-prose vs.\ formal-model) \emph{and} the decision procedure
(different solver, different input format); the residual shared
surface, stated rather than hidden, is the BTOR2 emission library ---
the site of incident I3, the platform's one recorded cross-route
common-mode failure \emph{through shared code} (shared misreadings are
the other kind, \S\ref{sec:bugs}) --- and the language-owned endpoint
interpreters.

\subsection{The compliance slice as a question set}
\label{sec:eval-bench}

\input{results/tables/bench}

The questions above were authored; \Cref{tab:bench} scales the same
discipline over a \emph{derived} question set nobody hand-picked. The
platform's RISC-V compliance slice --- self-checking programs in the
upstream riscv-tests HTIF \texttt{tohost} convention, built with the
pinned cross-toolchain at the upstream link base \texttt{0x8000\_0000}
--- doubles as a reachability benchmark: each program is reference-run
to completion, and for each data register the harness derives one
question whose target value the register demonstrably held (ground
truth \textsc{reachable}) and one whose target no register held at any
step (ground truth \textsc{unreachable} within the bound), with the
unrolling bound covering the whole run. Every question is then decided
independently along \emph{both} RISC-V routes. Three honesty notes.
First, the slice is curated in the upstream convention rather than
taken binary-for-binary from upstream: the stock \texttt{-p-} binaries
open with machine-mode CSR and trap setup that is outside the declared
user-ISA scope, so the slice re-creates the grading convention over
the user subset (the gap is typed, not hidden). Second, the ground
truth comes from the shared reference interpreter --- but that
interpreter is exactly the component the adequacy campaign validates
step-for-step against the Sail-generated gold simulator \emph{on these
same programs}, so the benchmark's ground truth is anchored outside
the platform. Third, the benchmark's discriminating power is scoped:
ground truth and both routes share the reference interpreter, so it
measures translator- and route-level defects --- a shared-misreading
defect of the MUL/ADD class (\S\ref{sec:bugs}) would pass it
undetected, and the external Sail differential, not this benchmark,
is what polices that class. The result: 78 questions over 10 programs (constant
dataflow, logic, shifts, 32-bit word ops, taken and untaken branches,
computed jumps, the store/load family, multiplication, division with
its RISC-V-defined edge cases, and compressed forms), both routes
agreeing on all 78 and all 78 matching ground truth, in under a minute
end to end --- with every one of the 39 bounded-unreachable verdicts
additionally corroborated by a full-bound replay of the direct route's
system through the strict BTOR2 interpreter (\S\ref{sec:eval-branch}'s
surrogate, recorded per question). Preparing this benchmark also forced one more fidelity
repair the base-0 probes could not see: the Sail-route front rebased
images to address 0, which silently breaks every pc-relative
absolute-address computation (\texttt{AUIPC}/\texttt{LA}) on images
linked at the upstream base --- found while squaring the translator
against the slice's link discipline, and fixed by carrying absolute
addresses through the Sail object (a versioned bump, with the squares
and this benchmark as regression).

\subsection{End-to-end case studies}
\label{sec:eval-cases}

\input{results/tables/cases}

\begin{figure}
\centering
\begin{tikzpicture}[x=1cm, y=1cm]
\node[art]  (p)      at (0.9, 0)    {source $p(x)$};
\node[art]  (z)      at (6.3, 0)    {artifact at $Z$};
\node[tool] (solver) at (6.3, -1.1) {solver};
\node[art]  (model)  at (6.3, -2.2) {witness $x_0$};
\node[tool] (ia)     at (0.9, -2.2) {$I_A$ replays $p(x_0)$};
\node[art]  (v)      at (3.6, -3.2)
  {$\varphi$ observed: \textsc{reachable}, $\tcb = \{I_A$ adequacy$\}$};
\draw[chk] (p) -- node[elbl, above] {$T_R$ (whole route)} (z);
\draw[chk] (z) -- (solver);
\draw[chk] (solver) -- (model);
\draw[chk] (model) -- node[elbl, above] {$\carry_R$ (carry-backs)} (ia);
\draw[chk] (ia) -- (v);
\end{tikzpicture}
\caption{The round trip every case study in \Cref{tab:cases} completes
(\Cref{thm:existential}): the witness comes back through the route's
carry-backs and is replayed by the source interpreter. Everything on
the way down is a \emph{discovery} device --- a defect in any
translator, hub, or the solver can lose the witness, never fake the
replay.}
\label{fig:replay}
\end{figure}

\Cref{tab:cases} instantiates \Cref{thm:existential} once per hub and
front-end family, completing the round trip of \Cref{fig:replay} at
the source each time. (1)~The \emph{C spine}: both routes answer
\textsc{reachable} for \texttt{a0}${=}47$, and the source-level replay
runs the compiled program in the shared RISC-V interpreter to the halt
in \texttt{\_start()} with \texttt{a0}${=}47$ observed --- the verdict
survives even if every translator and the solver were wrong.
(2)~\emph{Python}: ``is the trailing assert violable?'' for a function
with a bounded loop and a branch; Z3 finds the witness $x{=}4$, the
shared \textsc{QF\_LIA} evaluator independently re-checks the model
(\texttt{smt\_model\_ok}), and the input is replayed through the pinned
CPython down the taken branches, firing the assert. (3)~\emph{EVM}:
$6\times7{=}42$ on the operand stack, decided \emph{natively} (btormc
on the BTOR2 system, witness replayed by the strict BTOR2 interpreter)
and \emph{bridged} (through the SMT-LIB hub with Z3) --- solve-step
corroboration across two engine families, agreeing. (4)~\emph{CRN}: a
two-firing reachability question in a reaction network, witness
schedule replayed by the Petri-net interpreter. In all four, the final
trust rests on the source interpreter (plus the small fixed replay
harness \Cref{thm:existential} names), per the last row of the
ledger in \S\ref{sec:tcb}.

\subsection{The certified tier}
\label{sec:eval-proved}

\input{results/tables/proved}

\begin{figure}
\centering
\begin{tikzpicture}[x=1cm, y=1cm]
\node[lang] (b)    at (0.5, 0)    {btor2};
\node[hub]  (smt)  at (2.2, 0)    {smtlib};
\node[tcb]  (bb)   at (4.5, 0)    {bitwuzla BV$\to$CNF};
\node[art]  (cnf)  at (6.9, 0)    {CNF};
\node[tool] (cad)  at (6.7, -1.1) {cadical};
\node[art]  (drat) at (5.0, -1.1) {DRAT};
\node[tool] (dt)   at (2.9, -1.1) {drat-trim (untrusted)};
\node[art]  (lrat) at (0.7, -1.1) {LRAT};
\node[tcb]  (clpr) at (1.6, -2.2) {cake\_lpr (verified)};
\node[art]  (ok)   at (5.6, -2.2) {\texttt{s VERIFIED UNSAT}};
\draw[pred] (b) -- (smt);
\draw[chk] (smt) -- (bb);
\draw[chk] (bb) -- (cnf);
\draw[chk] (cnf) -- (cad);
\draw[chk] (cad) -- (drat);
\draw[chk] (drat) -- (dt);
\draw[chk] (dt) -- (lrat);
\draw[chk] (lrat) -- (clpr);
\draw[chk] (clpr) -- (ok);
\end{tikzpicture}
\caption{The pipeline \Cref{tab:proved} evaluates: the query is
bit-blasted to CNF, cadical's DRAT refutation is elaborated by
\texttt{drat-trim} --- an untrusted step that can only fail the
re-check, never fake a pass --- and \texttt{cake\_lpr}, the formally
verified checker, re-validates the LRAT against the CNF from scratch.
Shaded is what stays in the verdict's \tcb: the bit-blast and the
verified checker; the three agreeing engines and the elaborator are
discovery devices.}
\label{fig:cert}
\end{figure}

\Cref{tab:proved} inhabits the ledger's certified row
(\S\ref{sec:tcb}), twice --- via the pipeline of \Cref{fig:cert} --- once at each end of the difficulty
scale. Both questions are pair-derived and input-driven: eBPF programs
whose \texttt{CALL} models helper returns as free inputs. The first
squares one input and asks whether $x^2 = 3$; no square exists modulo
$2^{64}$, so the claim is universal over all inputs --- and its
refutation is essentially unit propagation on the low bits (the
elaborated LRAT is 18 bytes). The second masks and offsets two inputs
into $[2, 2^{16}{+}1]$, multiplies them, and asks whether the product
can be $2^{31}-1$: the Mersenne prime admits no factorization with
both factors in range, and refuting a $16{\times}16$-bit
multiplication is certificate checking at scale --- a 2.8\,MB DRAT
elaborating to an 11.8\,MB LRAT, re-validated in seconds. In both,
three engines corroborate \textsc{unsat}, bitwuzla bit-blasts the
query to CNF, cadical emits a DRAT refutation, \texttt{drat-trim}
elaborates it to LRAT --- an \emph{untrusted} step: a wrong
elaboration can only fail the re-check, never fake a pass --- and
\texttt{cake\_lpr}, the \emph{formally verified} CakeML LRAT checker
(soundness machine-proved down to the binary; here built natively
from its compiler-generated ARMv8 assembly), re-validates the proof
against the CNF from scratch.
The verdicts' \tcb\ is exactly the ledger's certified row: the
bit-blaster's BV$\to$CNF step and a verified checker, with all three
deciding engines and the elaborator demoted to discovery devices ---
the solve step's residue; the route hops that produced the artifact
remain accounted per the ledger. Both certified systems are also
corroborated below the whole certificate pipeline: fifty seeded random
helper streams replayed through the strict BTOR2 interpreter fire no
\texttt{bad} (recorded in the run data). The
reachable siblings ($x^2 = 4$; $x{\cdot}y = 46341^2$) correctly
decline to certify.

Wiring this exhibit produced one more architecture-caught defect,
live: the checker \emph{adapter} matched the substring
\texttt{VERIFIED} in the checker's output --- and \texttt{drat-trim}
reports failure as \texttt{s NOT VERIFIED}, which contains it. Every
check, sound or bogus, reported success. The bug could not surface on
hosts without the checker (the call raised before parsing) nor on real
certificates (which do verify); it fell to the \emph{negative
control} --- a bogus refutation of a satisfiable CNF that must fail
and did not. The fix parses the exact status line, and the control is
now a permanent test. This is the paper's own instrument audited by
the discipline it describes (cf.\ \S\ref{sec:bugs}), and a caution we
commend to anyone wiring a proof checker: \emph{a checker adapter
without a negative control is itself unchecked}. The verified
checker's adapter was written under that rule from the start --- wisely,
since \texttt{cake\_lpr} exits 0 even when checking fails; only the
exact \texttt{s VERIFIED UNSAT} line signals success, and both
checkers' negative controls are permanent tests. One honest subtlety
the exercise surfaced: for this instance the refutation is
propagation-dominated, so mutating single proof lines is not a valid
negative control (the checker legitimately completes the derivation);
the satisfiable-formula control is the sound one, since no valid
refutation of a satisfiable formula exists.

\subsection{Defects the architecture caught}
\label{sec:bugs}

\input{results/tables/bugs}

\paragraph{Two incidents in full}
The catalog's two most argument-bearing entries deserve their
narratives. First, the \emph{blind spot}: the RISC-V interpreter
\emph{and} the \texttt{riscv-btor2} translator mis-decoded
\texttt{MUL} (\texttt{funct7}${}=$\texttt{0x01}) as \texttt{ADD} ---
identically. Both legs of every square agreed, so the square was
structurally blind, and the defect was found by manual audit during
widening, not by the architecture. This is the residual failure mode
the calculus itself predicts --- \Cref{lem:agree}'s ``both unfaithful
with identical projected error'' --- landing not between two
translators but between a translator and the component
\Cref{thm:existential} crowns as the sole residual \tcb. Both sides
now hard-abort on unknown \texttt{funct7}, and the structural answer
is \emph{external} anchoring: the decoder is validated against 610
Sail-emitted instruction words and against an independently written
RV64C decompressor, construct by construct --- and the common-mode
round of \S\ref{sec:eval-escape} now measures this incident's whole
class directly. Second, the \emph{common
mode}: the shared BTOR2 emission library declared states before their
constant \texttt{init} values, so every emitted \texttt{init} line
was malformed --- in every BTOR2-emitting pair at once, both branches
of every RISC-V question included --- and the one solver then wired
(Z3, through the bridge) tolerated the malformation. Adding the
second and third engines turned it into hard rejections: solve-step
corroboration, not the squares, caught it (incident I3). Between them
the two incidents mark the exact boundary of \Cref{asm:diversity}:
shared code and shared misreadings are what agreement cannot see,
which is why the ledger keeps the emission library and the endpoint
interpreters inside every branch verdict's \tcb\ --- and why the
disjoint stacks of \Cref{fig:branch} and the external differentials
exist.

\subsection{An escape-rate estimate for the gate}
\label{sec:eval-escape}

\input{results/tables/escape}

\input{results/tables/escape_fig}

The catalog above is mined from history and has no denominator; the
MUL/ADD blind spot proves the gate's escape rate is nonzero.
\Cref{tab:escape} measures it (\Cref{fig:escape} draws the gate
stack): seeded semantic mutations of the
\texttt{riscv-btor2} translator's emissions --- uniform rules modeling
systematic mis-lowerings (every \texttt{sext} emitted as \texttt{uext},
incident I2's family) and site rules modeling single-site defects
(incident I1's family); operand swaps only on non-commutative
operators, so no trivially-equivalent mutants; rules that change no
probe artifact excluded from the denominator --- each run through the
architecture's gates in the order the platform applies them. The
experiment earned its keep before its final numbers did: in its first
round, both \texttt{srl}$\to$\texttt{sra} mutants escaped \emph{all
three} gates (36 of 51 caught at the square layer), because every
ALU probe ran its construct on degenerate all-zero operands, on which
logical and arithmetic shifts agree --- the ``96/96 conjoined'' verdict
was faithfulness \emph{on the probes}, and the probes underdetermined
the constructs. Hardening the probes with mixed-sign operands moved 50
of 53 catches to the square layer but let \texttt{ult}$\to$\texttt{ulte}
escape --- strictness is observable only at \emph{equal} operands ---
so the probes now run an equal-operand and a mixed-sign instance per
construct (incident I23; the ratchet kept every previously covered
construct covered throughout). In the final round, 51 of 55 applicable
mutants die at the square suite, the remaining 4 at the
compliance-derived benchmark, and none escape. Three honest notes:
zero escapes means this \emph{mutation family} is now covered, not
that the escape rate is zero;
the authored branch questions caught nothing in any round (the derived
benchmark subsumes them as a differentiator, a finding in itself); and
the denominator is one translator's emission space, not the platform.

\paragraph{The common-mode round}

\input{results/tables/common}

Single-leg mutation cannot model the MUL/ADD class where both legs of
the square are wrong \emph{identically} --- so a second experiment
injects exactly that (\Cref{tab:common}): six shared misreadings,
each applied as the same uniqueness-checked source patch to \emph{both}
the reference interpreter and the \texttt{riscv-btor2} translator
(in-memory shadows; the repository is untouched). The structural
predictions hold, measured: the square suite is blind on all six (both
legs agree on every probe), and so is expected-based grading in the
\emph{poisoned world}, where the benchmark's ground truth is derived
by the mutated interpreter itself. What catches the mutants are the
gates \emph{outside} the shared misreading ---
\Cref{fig:escape}'s stack read as ordered by independence, each gate
derived farther from the shared code than the last: four die at
cross-route
disagreement with the Sail-derived branch, and two survive even that
--- their misreadings never surface in the per-register values the
derived questions sample --- falling only to the external
Sail-simulator differential, which (run as a parallel column) catches
all six. The round also caught its own instrument, live: in its first
run, \texttt{slt-as-sltu} \emph{escaped every gate} --- the slice's
compare vectors were all same-sign, on which signed and unsigned
comparison agree, so neither the derived questions nor the
differential's executed streams could see the difference (incident
I24, the slice-level instance of I23's probe lesson, and the
platform's first measured full-stack escape). The ratchet's answer
was upstream-faithful --- \texttt{rv64ui-slt}'s mixed-sign and
equal-operand vectors added to the slice --- after which the round
completes with zero escapes. The honest reading mirrors
\Cref{asm:diversity}'s scope exactly: against shared misreadings every
gate built from the shared reading is blind \emph{by construction},
corroboration is only as
strong as the most independently derived anchor, and that
anchor's vectors are themselves an instrument the discipline must
audit.

\subsection{The other direction: an LLM plays the platform}
\label{sec:eval-player}

\input{results/tables/player}

\begin{figure}
\centering
\begin{tikzpicture}[x=1cm, y=1cm]
\node[art] (q) at (3.6, 0)
  {reachability question + platform-established ground truth};
\node[tool, align=center] (aa) at (1.4, -1.2)
  {arm A: unaided LLM\\(question text only, no tools)};
\node[tool, align=center] (ab) at (5.9, -1.2)
  {arm B: LLM player\\(same question + repository)};
\node[hub, align=center] (plat) at (5.7, -2.4)
  {\hg: routes, squares,\\solvers, checkers};
\node[art, align=center] (va) at (1.4, -3.5)
  {verdict\\(say-so, high confidence)};
\node[art, align=center] (vb) at (5.9, -3.5)
  {verdict + machine-checked\\evidence artifact};
\node[art] (g) at (3.6, -4.5) {graded against ground truth};
\draw[chk] (q) -- (aa);
\draw[chk] (q) -- (ab);
\draw[chk] (ab) -- (plat);
\draw[chk] (plat) -- (vb);
\draw[chk] (aa) -- (va);
\draw[chk] (va) -- (g);
\draw[chk] (vb) -- (g);
\end{tikzpicture}
\caption{The two-arm player protocol of \Cref{tab:player}: one fresh,
context-free agent per question per arm, no retries. Both arms are
graded against the same platform-established ground truth; what
differs by construction is what the verdict \emph{rests on} --- arm B
can take no unchecked step, so every answer carries the evidence
artifact of \Cref{tab:player}'s final column.}
\label{fig:player}
\end{figure}

The introduction's two-directional experiment has, until here, reported
only its build half. This subsection is a first, deliberately small
evaluation of the \emph{player} half: 12 reachability questions with
platform-established ground truth (6 reachable / 6 unreachable, spanning
RISC-V signedness and load-width traps, eBPF square-residue and
bounded-factorization questions, and Python assert-violability), posed
to fresh, context-free frontier-LLM agents under two protocols
(\Cref{fig:player}). Arm A
(unaided) received only the question text, instructed to answer by
reasoning with no tools. Arm B (player) received the same question plus
the repository and the head-construction snippet, instructed to decide
\emph{via the platform} and report the verdict with its evidence basis.
One agent per question per arm, no retries; \Cref{tab:player} lists the
questions and both arms' verdicts with each arm-B answer's evidence
artifact; prompts, ground truths,
per-run outputs, and grading are in the artifact
(\texttt{results/llm\_player/}).

Both arms scored 12/12. That is the honest headline, and it reframes
what the platform buys at this question scale: not raw accuracy ---
the unaided frontier model simulated the loops, caught both signedness
traps, and factored 2147766287 by Fermat's method in its head --- but
the \emph{evidence class} of the answer. Every arm-A verdict rests on
the model's say-so (all were reported at high confidence, so
confidence would not have flagged a wrong one); every arm-B verdict
carries a machine-checked artifact: two-route agreement on the RISC-V
questions, hedged explicitly as bounded-$k$ verdicts resting on route
fidelity; witness replays through the strict interpreters on the
reachable eBPF/Python questions; and tier-raises to \tprov\ on the
unreachable eBPF questions. The sharpest contrast is the
bounded-factorization question: arm A answered correctly by
\emph{recalling} that 1073741789 is the largest prime below $2^{30}$
--- opaque memory, exactly the evidence the calculus cannot grade ---
while arm B produced a 17\,MB LRAT certificate re-validated by the
formally verified checker, with the trusted base named. One player
even devised its own positive control unprompted (flipping the load
question's target to the sign-extended value and confirming
\textsc{reachable} on both routes). Four limitations, stated plainly:
the protocol scripted which platform entry points to use (the players
executed and interpreted; they did not discover the platform cold);
the subject model is from the same family that built the platform;
the grading is partly circular --- ground truth is
platform-established and arm B answers \emph{via} the platform, so arm
B can fail essentially only by mis-operating or misreading the tools;
and
12 questions at this size cannot separate the arms on correctness ---
questions hard enough that unaided reasoning actually fails are the
obvious next step.

\subsection{Performance, determinism, and scale}
\label{sec:eval-perf}

\input{results/tables/perf}

\input{results/tables/scale}

Translation is not the cost center\ifarxiv{} (\Cref{tab:perf})\fi: whole-route translation of the loop
question ($k{=}25$ unrolling, $\sim$200\,kB of SMT-LIB) is
$\sim$5\,ms cold and a sub-millisecond content-addressed cache hit warm
(\Cref{prop:cache} cashed in), against a Z3 decide in the tens of
milliseconds on either
route; the
twice-and-diff determinism check passes on both routes. The numbers say
the honest overhead of the whole discipline --- checking every hop,
re-running for determinism, replaying witnesses --- is small against
the solver call it wraps. \Cref{tab:scale} probes where the pipeline
stands as questions grow, on two axes: unrolling the summation loop to
$N{=}100$ ($k{=}505$, a 4\,MB artifact) keeps whole-route translation
under 0.1\,s and the decide under 1\,s, growing roughly linearly; and
the certified tier survives the bounded-factorization family from
propagation-size to search-size --- certificates growing from 25\,kB to
an 11.8\,MB LRAT, each re-validated by the verified checker in seconds.
Neither axis reaches industrial scale, but both cross the region where
``toy'' criticisms live: the pipeline's costs grow smoothly, with no
cliff through the measured range.

\ifarxiv
\subsection{Post-snapshot exhibits}
\label{sec:eval-exhibits}

Every number above is the dated snapshot, regenerable from the artifact
by one script; nothing built since is mixed into those tables. What the
platform has grown since is reported here as \emph{exhibits} ---
artifact-checkable demonstrations, each a named test or command, not
benchmark claims:

\begin{itemize}
\item \emph{Constraint enforcement, corroborated both ways.} The shared
  BTOR2 evaluator enforces environment constraints (per-row
  observables; a violating row ends the run), and wiring it surfaced a
  latent bridge defect of independent interest: asserting constraints
  globally over the unrolling masks a bad reached on a valid prefix
  before a later, inevitable violation --- and disagrees with the
  native checkers on exactly that system. With the per-frame encoding,
  the evaluator, the bridged engine, and the native checker agree on
  the constrained corpus in both verdict polarities.
\item \emph{The books in CI.} The capped route-grader runs with the
  ledger configured; cost profiles accumulate per runner class; and the
  annotated route report reproduces, on runner-measured translate
  totals, the expected dominance mark --- the Sail-mediated
  RISC-V route marked against the direct one at equal assurance and
  direction, both still listed.
\item \emph{Independence, judged from declarations.} The trust advisor
  finds both ISA branches independent from their declared semantic
  artifacts (prose manual vs.\ Sail model), and, on a single-route
  source (Wasm), returns the honest demand --- an independent pair from
  a new artifact --- rather than a corroboration claim.
\item \emph{The demand board} of \S\ref{sec:books}, and the
  spurious-counterexample refinement demonstration in the
  \texttt{btor2-havoc} tests.
\end{itemize}

These exhibits are the seeds of the next benchmark families --- HWMCC
ingestion for the hub and its abstraction pairs, a cost-calibration
corpus for the ledger's profiles, and question-campaign corpora for the
demand books --- measurements we intend to run, not only narrate.
\fi
