%% §8 — written out in Phase 4.

\section{Limitations and conclusion}
\label{sec:conclusion}

Four walls stand at every registry size, and the theorems are drawn
up to them, not past them. \emph{Computability}: a route re-expresses
a question, never changes its degree; unbounded universal questions
rest on bounded claims under declared bounds. \emph{The adequacy
floor}: no pair can make a question about something the source
semantics does not define into a question about the program.
\emph{Cost}: deciding is exponential in the worst case everywhere
that matters, budgets are permanent verdict classes, and a plateau in
the curve is a finding — the measured practice-frontier of the domain
— not a failure. \emph{Anchor supply}: capability scales with
generated pairs; trust scales with independent formalizations of real
semantics, which exist in small finite supply. Within the walls, the
claim of this paper is deliberately modest in kind and, we believe,
useful in consequence: an architecture whose exploration is sound,
monotone, self-directing, relatively complete, and terminating per
benchmark — provably, with the assumptions priced — turns ``we don't
know how to decide this'' from a shrug into a typed, clustered,
cost-profiled, evidence-citing specification of the missing
instrument.

The maps compound, because pairs are shared: every benchmark starts
on the ground the last one won. The near-term work is scale — widening
the pinned corpus family by family, the software-verification spine
as the second act — and the mandate rung, so that saturation's
mechanical middle runs at machine speed between two human acts:
writing the mandate, and reading the map. Two extensions are staged
beyond it, at deliberately different maturities, both demand-driven.
The synthesis lane — the books demanding decision procedures, a gate
falsifying candidates, synthesized engines gated harder than the
binaries they would join — is tooled and shadow-first; whether it
moves real frontiers is a finding the saturation reports will or
will not deliver, and we claim nothing sooner. And on the trust
axis, demanding \emph{mechanical proofs} of a pair's square through
the same books — a certificate demand priced against an independent
second route, translation validation per run or refinement proofs
where a mechanized semantics exists — is designed, not built. The
vision is the means. The map is the end.
