%% §6 — written out in Phase 4: the kit frozen, the design oracle, the
%% atlas (O1), challenge bundles, and the pre-registered protocol in
%% place of a benchmarks section. No measurements, by design.

\section{The domain kit, and the far side as a design oracle}
\label{sec:kit}

\subsection{The kit}
\label{sec:kit-list}

Pointing the explorer at a domain costs exactly four things, and the
theorems of \Cref{sec:theorems} are conditional on them — this list is
the instantiation obligation of F6, frozen:

\begin{enumerate}
\item \textbf{A reasoning hub}: at least one language a mechanized
  decision procedure consumes directly, with declared question shapes
  and declared budgets. (Without it, no question clears the shape and
  cost conditions.)
\item \textbf{Deterministic shared interpreters} for every language
  touched, exposing named observables. (Without them, squares cannot
  be checked and F1's replay proves nothing.)
\item \textbf{At least one external semantic anchor}, if trust beyond
  per-run checking is wanted. (Without it, the trust axis saturates
  at \textit{checked}; the theorems still hold, and the map says so.)
\item \textbf{A pinned benchmark}: a finite question set with recorded
  provenance. (Without the pin, F5's fixpoint is not defined.)
\end{enumerate}

The kit is deliberately mundane: hardware model checking supplies all
four today; C program analysis supplies them down a longer spine; and
the platform's own field-blind corners (chemical reaction networks,
molecular notation) are the cheap existence proof that nothing in the
list assumes computation-adjacent
languages~\citep{hurdygurdy2026calculus}.

\subsection{The far side as a design oracle}
\label{sec:kit-oracle}

A terminal board is not a failure list. By the honest-failure
discipline, every open question carries typed evidence, and each
obstacle class comes with an \emph{extraction operator} turning that
evidence into the specification of the missing instrument:

\begin{center}
\small
\begin{tabular}{@{}lll@{}}
\toprule
obstacle & evidence on the books & extraction operator \\
\midrule
shape & the blocked $\varphi$s verbatim & fragment hull + atlas location \\
shape, charted family & the hull + the solved census & procedure-synthesis brief \\
cost & measured curves at the wall & binding-parameter fit \\
cost, via abstraction & the spurious corpus & separating predicates \\
trust & uncorroborated verdicts & certificate and anchor joins \\
outside the closure & the wall itself & nearest in-closure relaxation \\
\bottomrule
\end{tabular}
\end{center}

Four operators are already mechanical. The \emph{binding-parameter
fit} is the report's failure-mode reading: blocked instances get
their cost curve measured by an ascending probe, and the fit
classifies — exponential in the unrolling bound means deeper search
will not close it (demand an unbounded engine or a property
transformation); linear means depth is affordable (raise the budget
before demanding an instrument). The \emph{atlas location} is a
registry-side reference table of the known decidability landscape:
a shape-blocked demand arrives locating itself — the setting in which
its shape is decidable, the procedure family that decides it
natively, and the \emph{known crossing}, the classical reduction that
would discharge it with an endo-pair on an existing hub instead of a
new logic (liveness-to-safety for liveness; self-composition for
2-safety hyperproperties; monitor weaving for temporal shapes). The
atlas is reference data, protected like every other instrument the
gate trusts, and a shape it does not know reads \textit{uncharted},
never a guess — and it is load-bearing: chartedness is what
classifies a procedure demand as instantiation (inside the known
set, blocking saturation) or discovery (outside, on the frontier).
The \emph{scout} measures before anyone designs: candidate
embeddings for a blocked shape are prototyped and their encoding
blowup measured, so a procedure demand arrives justified by numbers
rather than taste. And the \emph{procedure-synthesis brief} turns a
charted, procedure-shaped demand into a self-contained work item —
the fragment hull of its citing $\varphi$s, the atlas location and
family, the required contract, the census slice that will falsify
the build, and the certificate obligation. The recommender is
audited in kind: every recommendation is a prediction of closure,
checked ex post against realized closure. The remaining operator
(separating predicates mined from the persistent spurious corpus) is
named work, staged behind the same shadow discipline as every other
delegated judgment.

\paragraph{Challenge bundles}
Frontier pairs are exportable: a board entry carries pinned citing
questions, declared budgets, a required contract, and the near side's
baseline census — a self-contained challenge for the domain's solver
community. This is the sense in which the map's far side is a demand
oracle for the ecosystem: competitions supply benchmarks to the
platform; terminal boards supply open problems, with evidence counts,
back.

\subsection{The pre-registered protocol}
\label{sec:kit-protocol}

This paper deliberately has no benchmarks section: its place is the
saturation report the following protocol will produce, and
pre-registering the protocol is what keeps that report honest.

\textbf{The benchmark.} The hardware model-checking competition
corpus (HWMCC~\citep{niemetz2018btor2}): designed by others, labeled,
native to the platform's bit-level hub — no translation debt to pay
before the loop starts. Ingestion is streamed-with-pin from a
community mirror at a fixed commit, every instance sha256-pinned; the
campaign starts from the already-pinned slice and widens by families,
one pinned extension at a time. The natural second act is software
verification (SV-COMP~\citep{beyer2021svcomp}) down the full
C-to-solver spine.

\textbf{The protocol.} (1) Pin the suite; establish ground truth
where labels exist. (2) Run the loop's player over every question,
books on, \textit{origin=campaign}, suite-tagged — manufactured
demand is displayed apart from organic demand by construction.
(3) What answers, answers with its way-census; what does not lands on
the books with its first failing obstacle, and blocked instances get
their cost curves measured. (4) Derive the board; promote what a
human (or, when earned, a mandate) registers; build, gate, merge.
(5) Re-run. The ratchet makes the curve monotone; stop at the
fixpoint, which is an exit code.

\textbf{Declared caps.} Unrolling bound, wall-clock and memory per
decide, instance count per iteration, one instance in memory at a
time; all caps ride in every result's provenance, and capped results
are labeled capped, never implied full-suite.

\textbf{Pre-registered honesty.} The curve may plateau on cost; the
plateau is a finding — the measured practice-frontier of the domain —
not a failure. \textit{unknown} and \textit{resource-out} are
counted, never hidden. A terminal board dominated by frontier-only
entries is a \emph{successful} saturation. The tooling (the loop
driver, the saturation checker, the report generator, the promotion
command) is built, tested, and has been exercised end to end on the
pinned slice; no number from any run appears in this paper.
