%% §4 — written out in Phase 4: the explorer's engine, compressed to
%% what §5's theorems consume.

\section{The loop}
\label{sec:loop}

\begin{figure}
\centering
\begin{tikzpicture}[font=\footnotesize, >=Stealth,
  box/.style={draw, rounded corners=1pt, align=center,
              inner xsep=4pt, inner ysep=3pt, minimum width=9.5em},
  lab/.style={font=\scriptsize\itshape, align=center}]
  \node[box] (ask)   {the player plays\\every $q \in B$};
  \node[box, right=9mm of ask] (reg) {the registry, grown\\(the ratchet holds)};
  \node[box, below=8.5mm of ask] (books) {the books: obstacle,\\target, origin, suite};
  \node[box, below=8.5mm of reg] (gate) {the gate + negative\\controls (solver gate\\for procedures)};
  \node[box, below=8.5mm of books] (board) {frontier pairs, derived:\\required contracts};
  \node[box, below=8.5mm of gate] (lanes) {three lanes: PRs $\cdot$\\builders $\cdot$ procedures};
  \node[box, very thick, below right=7mm and -8.5em of board]
    (valve) {the valve: a human registers (a mandate instantiates)};
  \draw[->, dashed] (ask.north) -- +(0,5mm)
    node[above, lab] {answered: way-census, into the map};
  \draw[->] (ask) -- node[lab, left] {blocked:\\why-not} (books);
  \draw[->] (books) -- node[lab, left] {group by\\target} (board);
  \draw[->] (board.south) ++(0,-0.2mm) |- ([yshift=1.5mm]valve.west)
    node[pos=0.8, lab, above] {promote};
  \draw[->] ([yshift=1.5mm]valve.east) -| node[pos=0.2, lab, above] {briefs}
    (lanes.south);
  \draw[->] (lanes) -- (gate);
  \draw[->] (gate) -- node[lab, right] {admitted} (reg);
  \draw[->] (reg) -- node[lab, above] {re-ask} (ask);
\end{tikzpicture}
\caption{One loop iteration. Demand descends on the left --- every
failure a typed record on the books, aggregating into frontier pairs
--- and supply ascends on the right, but only through the valve at
the bottom: registration is the one human act (or its earned,
revocable mandate). The two planes meet nowhere else: answers never
write, growth never answers. Monotonicity (\Cref{thm:monotone})
makes the cycle a ratchet, never a treadmill.}
\label{fig:loop}
\end{figure}

The explorer runs two loops with different payoffs
(\Cref{fig:loop}). The
\emph{capability} loop is triggered by unanswerable questions: the
diagnosis (\Cref{def:diagnosis}) names the first failing condition
and, with it, a generation target — a missing pair, a wider
projection, a missing reasoning language, a reduction, an independent
anchor. The \emph{trust} loop is triggered by answers weaker than the
asker's floor: a second route answers no new question at all, but an
\emph{independently derived} second route manufactures fidelity, and
the trust advisor names the honest option — an existing independent
branch, a route from a new semantic artifact, or saturation of the
anchor supply, stated as such. Both loops consume the same books and
produce the same currency: demand records that aggregate, per
generation target, into the frontier objects of \Cref{def:tiers}.

\paragraph{Recommended, then registered}
A pair must pay for itself by removing a named obstacle, so a pair is
\emph{recommended by its evidence} before it is registered: a brief
opens by citing the demand behind it — mechanically pre-filled from
the frontier object it promotes, required contract and citing
questions verbatim. Registration itself stays a human act, the
platform's one scope valve: the books recommend, volume is not a
verdict, and a human decides what enters the queue. The delegated
generalization is the \emph{mandate}: a human registers not a pair
but a region — a benchmark, the obstacle classes in scope, the
admissible languages, the protected floors — within which
demand-cited briefs may be registered mechanically, earned by the
same shadow discipline as autonomous merging and revoked by the hand
that wrote it. The valve does not move; what is delegated is its
mechanical instantiation.

\paragraph{The two production lanes}
Pairs are produced by others, through pull requests the gate checks
exactly as it checks a builder agent's — green CI is designed to
\emph{mean} safe-to-merge — and by the platform's own builders,
dispatched at partial pairs and unregistered briefs, integrated by a
coordinator whose merge queue orders, fans out re-validation, and
proposes; its autonomy is graduated rung by rung on a ledger of shadow
trials, never granted once~\citep{hurdygurdy2026calculus}. Both lanes
feed the same registry under the same ratchet, which is what makes F2
a property of the system rather than of one workflow.

\paragraph{The third lane: decision procedures}
The books can also demand what no route can supply --- a decision
procedure itself. The demand arises in exactly two places: a blocked
shape the atlas locates in a known procedure family that no
registered hub declares, and a scouting run in which every
prototyped embedding measurably explodes, the blowup riding as the
demand's evidence. A candidate procedure, however authored, enters
through a \emph{solver brief} mirroring the pair brief --- declared
shapes, a budget schema, a per-shape-and-verdict certificate
obligation, and a \emph{lineage} declaration, so that corroboration
counts only agreement across disjoint lineages and a teacher's
student cannot launder agreement into independence --- and an
admission gate that is the pair gate's analogue: replay of the
corroborated solved region (any disagreement fails admission
outright; agreement banks corroboration for every replayed
question), canaries and verdict-flip mutants (a candidate that
cannot be made to fail is not checked but unfalsifiable),
certificate discipline, and budget honesty (\textit{resource-out} at
the gate is a recorded cost profile, not a failure). One inversion
is the point: a synthesized in-process procedure is gated strictly
\emph{harder} than the pinned binaries beside it --- sandboxed like
untrusted pair code and determinism-checked, with no
external-engine exemption --- so the least-trusted authors produce
the most-checkable artifacts. The lane sits deliberately behind the
valve: procedure design is a creative act, no mandate rung exists
for it, and under every mandate the demand escalates to a human.
Its tooling is exercised end to end --- the reference inhabitant is
a hand-built bounded-enumeration engine admitted through the gate
--- and no more is claimed here: whether synthesized procedures move
real frontiers is exactly a question the saturation reports are
designed to answer.

\paragraph{CEGAR is the cost axis's engine}
Against the cost condition, the one reduction that helps is the
abstraction pair, and the loop it induces is counterexample-guided
refinement with one structural difference from every CEGAR engine
built inside a solver: the refinements are not throwaway internal
state but registered, audited, reusable pairs. A spurious
counterexample — an abstraction's witness that fails source replay —
is a refinement demand on the books; the refined abstraction joins
the graph and cheapens the next thousand questions. The abstraction
advisor makes the dial mechanical (the question's cone of influence,
the zero-loss free set, a farthest-first refinement ladder), and the
report's failure-mode reading (\Cref{sec:kit}) says when refinement
is the wrong remedy altogether.

\paragraph{The driver, and the structural valve}
The loop's mechanical form is deliberately one iteration per
invocation: pin the benchmark, play every question (the platform's
own decide plumbing for hub-native questions; the general player
plugs in at the same seam), book every spent verdict as a cost demand
with an ascending probe so the blocked instance's cost curve is
measured, re-diagnose, derive the board, deposit one self-contained
iteration record, regenerate the report. The loop then \emph{stops}:
registration happens between invocations, by a human or a mandate,
and the next invocation measures the growth. The human valve is
structural, not a prompt — there is no code path from a campaign run
to a registry write.
