%% §2 — the frontier problem. Drafted in Phase 1, in lockstep with
%% mechanization/Calculus/Frontier.lean: every definition below is the
%% Lean definition, prose first. This section is the paper's
%% contribution of *problem* and must stand alone.

\section{The frontier problem}
\label{sec:problem}

\paragraph{Domains}
A \emph{domain} is where questions live and decisions happen. Its
signature supplies, and nothing else about it matters:
languages, each carrying a \emph{formal semantics} --- a definable
meaning function over its programs, exposing named \emph{observables}
--- with formal semantics the only admission criterion; among them
some \emph{reasoning languages}, each owning mechanized decision
procedures with declared budgets and, where available, certificate
checkers; and a supply of independent \emph{semantic anchors} ---
external formalizations against which a translation can be checked by
something its author did not write. Hardware model checking is a
domain; C program analysis is a domain; chemical reaction networks
are a domain. The theorems of \Cref{sec:theorems} use only the
signature, which is what \emph{in any given domain} will mean.

\paragraph{Questions}
A \emph{question} $q = (p, \varphi)$ is a program $p$ in one of the
domain's languages and a condition $\varphi$ over $p$'s observables,
asked existentially (is $\varphi$ reachable?) or universally (does
$\varphi$ hold for all inputs, within declared bounds?). A
\emph{benchmark} $B$ is a pinned, finite set of questions with
recorded provenance --- pinned, so that ``the benchmark'' names the
same set at every loop iteration.

\paragraph{Items, registries, candidates}
The unit of admission is the \emph{item}: either a \emph{pair} --- a
deterministic, contract-carrying translation edge between two
languages~\citep{hurdygurdy2026calculus} --- or a language-attached
\emph{capability}: a solver, a checker, an anchor. A \emph{registry}
$G$ is the set of admitted items; growth is additive (items are
admitted, never silently removed), and every admission is gated. A
\emph{candidate} answer for $q$ rests on finitely many items: the
route's hops plus the destination capabilities it invokes --- one
list. Pairs for edges, capabilities for nodes: a candidate is the
route-level object, and it is the only object the definitions below
need.

\begin{definition}[Answerability]
\label{def:answerable}
Fix, per domain, $N$ \emph{conditions} on candidates, checked in a
fixed \emph{diagnosis order}. \hg{} instantiates $N = 5$:
\emph{connectivity} (the candidate is a route from $p$'s language to
a reasoning language), \emph{loss} ($\varphi$'s observables survive
the route's composed projection), \emph{shape} ($\varphi$'s logical
form is one the destination's procedures decide), \emph{cost} (a
procedure returns a verdict other than
\textit{unknown}/\textit{resource-out} within its declared budget),
and \emph{trust} (the answer meets the asker's assurance floor).
Write $S_k(G, q)$ for the candidates available in $G$ that survive
the first $k$ conditions --- the \emph{answerability filtration},
shrinking as $k$ grows and growing as $G$ grows. Then $q$ is
\emph{answerable at $G$} iff $S_N(G, q) \neq \emptyset$.
\end{definition}

Two words in the title are fixed by this definition.
\emph{In practice}: the cost condition quantifies over declared
budgets and measured runs, never over asymptotics ---
\textit{resource-out} is a permanent verdict class, and a route
without a measurement reads \textit{unmeasured}, never cheap.
\emph{Reducible}: answerability quantifies over candidates built from
admitted items, so the answerable set is the closure of the domain's
questions under the admitted reductions --- not an oracle's reach.

\begin{definition}[Diagnosis]
\label{def:diagnosis}
For $q$ unanswerable at $G$, the \emph{diagnosis} is the first
failing condition: the $k$ with $S_k(G,q) \neq \emptyset =
S_{k+1}(G,q)$. \Cref{thm:gradient} shows it exists, is unique, and
under growth only advances --- ``the obstacle'' is honest vocabulary,
and it is the frontier's local gradient: which kind of instrument
would move the map \emph{here}.
\end{definition}

\paragraph{The currency: three tiers of pair}
A pair carries a \emph{contract}: its assurance class, the direction
of its square, its kept observables, its measured
cost~\citep{hurdygurdy2026calculus}. Contracts compose along a route
by the componentwise meet and are ordered by strength; the meet is
what makes the following duality compose.

\begin{definition}[Required and achieved contracts; the tiers]
\label{def:tiers}
An implemented pair carries an \emph{achieved} contract, verified
from above: evidence establishes \emph{at least this}. A
\emph{frontier pair} carries a \emph{required} contract, demanded
from below: the join, over the open questions that cite it, of the
observables they need kept, the direction and assurance they need,
with cost kept as the histogram of their budgets. It also carries its
\emph{design evidence} --- everything the books know about how to get
there (\Cref{sec:kit}) --- and, where the demanded target is a
reasoning language that does not exist, a \emph{hypothetical}
endpoint: a language sketch, a typed hole in the graph. The registry
thus holds one lifecycle in three tiers:
\[
\textit{frontier} \;\longrightarrow\; \textit{registered}
\;\longrightarrow\; \textit{partial} \rightarrow \textit{built},
\]
where \emph{frontier} means design unknown, \emph{registered} means a
human admitted a design (the brief's ``intended translator'' is
filled in), and \emph{partial}/\emph{built} mean an achieved,
measured contract. Registering a frontier pair \emph{means}
exhibiting a design whose achievable contract dominates the required
one; promotion only ever advances, and the evidence payload travels
(\Cref{thm:currency}).
\end{definition}

Demand is not only edge-shaped. A missing \emph{capability} --- a
decision procedure no registered hub supplies --- is demanded in the
same currency, and a curated atlas of the known decidability
landscape (\Cref{sec:kit}) draws its line against the known set: a
\emph{charted} procedure family (setting, family, and canonical
crossings known) lies inside --- instantiation, and it blocks
saturation honestly --- while an \emph{uncharted} shape lies outside,
labeled discovery, never a guess.

\begin{definition}[Conditional routes]
\label{def:conditional}
A route may mix implemented and frontier hops. Its \emph{conditional
contract} is the meet over all hops --- achieved contracts for
implemented hops, required contracts for frontier hops --- and it is
read as: \emph{if} each frontier hop is discharged by a pair
achieving its requirement, this route answers $q$ with at least this
contract. Conditional routes are never executable; they are the
plan-shaped reading of the map, and \Cref{thm:currency} is what makes
them sound.
\end{definition}

\begin{definition}[Saturation]
\label{def:saturation}
$B$ is \emph{saturated at $G$} when the pair-shaped demand its open
questions derive contains no \emph{registered-able} candidate over
the known set --- no tier-2 design over the registered languages,
solvers, and anchors would add a feasible way, a cheaper way, or a
more trusted way --- i.e., every question in $B$ is in one of two
terminal states: \emph{solved, all ways}, its answer carrying the
census of every feasible route with measured cost profile, assurance
class, and corroboration; or \emph{open, on the books}, its frontier
pairs (if any are honest to name) all lying outside the known set,
partitioned by obstacle. Saturation is an emptiness check on a
derived set --- mechanically detectable, no judgment call --- and it
is \emph{reopened only by good news}: a new solver, anchor, or logic
re-enters the known set exactly where the terminal board says it
plugs in.
\end{definition}

\paragraph{The map}
The deliverable of saturating $B$ is its \emph{map}: the answered
fraction of $B$ per loop iteration (monotone, by
\Cref{thm:monotone}); cost per answer across iterations; per solved
question, the way-census; and the \emph{terminal board} --- the
surviving frontier pairs with their obstacle partition, required
contracts, design evidence, and distinct-question counts. The board
is the domain's future-work section, machine-generated; the map is
drawn in the same currency the instrument is built from.
