%% §7 — written out in Phase 4.

\section{Related work}
\label{sec:related}

\paragraph{Decidability by reduction}
Answering a question by carrying it to a decision procedure is a
classical discipline. In logic it runs from quantifier
elimination~\citep{presburger1929,tarski1951decision} through
interpretations that transfer decidability between
theories~\citep{tarski1953undecidable} to reduction into expressive
decidable hubs --- Rabin's S2S, into which many decidable theories
interpret~\citep{rabin1969trees} --- and the composition
method~\citep{makowsky2004fefermanvaught}; algorithmic
meta-theorems~\citep{courcelle1990graph} and well-structured
transition systems~\citep{finkel2001wsts} decide whole
logically-delimited families at once. In verification practice the
same move is institutional: bounded model checking reduces to
SAT~\citep{biere1999bmc}, theory combination composes decision
procedures~\citep{nelson1979simplification}, intermediate
verification languages pipeline programs to
SMT~\citep{barnett2005boogie,filliatre2013why3}, and Ivy makes
landing in a decidable fragment the design objective
itself~\citep{padon2016ivy,padon2017epr}. Each strand fixes the
reduction --- one interpretation, one pipeline, one fragment --- and
asks what it decides. Answerability (\Cref{def:answerable}) makes the
same move an object of study: the answerable set is the closure of a
domain's questions under an evolving, gated registry of reductions,
budgeted cost sits inside the definition rather than after it, and
the deliverable is the boundary --- every open point carrying its
first failing obstacle and the instrument that would move it.

\paragraph{Single-edge theories}
Certified compilation~\citep{leroy2009compcert,kumar2014cakeml} and
translation validation~\citep{pnueli1998tv,necula2000tv,lopes2021alive2}
are statements about one translation; the instrument
paper~\citep{hurdygurdy2026calculus} develops the graph-level theory
this paper consumes. What is new here is neither the edges nor the
graph but the \emph{use}: the graph run as an explorer whose product
is the boundary itself.

\paragraph{Abstraction and CEGAR}
Abstract interpretation~\citep{cousot1977abstract} supplies the lax
half of the directional square, and counterexample-guided
refinement~\citep{clarke2000cegar,kurshan1994computer} the cost
loop's shape. The structural difference is bookkeeping with
consequences: every refinement is a registered, audited, reusable
pair, and every spurious counterexample persists on the books as
evidence for the next abstraction's required contract, rather than
dying inside one solver run.

\paragraph{Certifying answers}
Proof-carrying code~\citep{necula1997pcc}, certifying
algorithms~\citep{mcconnell2011certifying}, verifiable
witnesses~\citep{beyer2015witnesses}, and checked proof
formats~\citep{wetzler2014drat,tan2021cakelpr,andreotti2023carcara}
supply the platform's trust vocabulary. The frontier reading adds
that the \emph{absence} of a certificate is itself data: trust-blocked
questions aggregate into capability demands naming the missing
checker or anchor.

\paragraph{Competitions and infrastructure}
HWMCC~\citep{niemetz2018btor2}, SV-COMP~\citep{beyer2021svcomp}, and
execution infrastructure like StarExec~\citep{stump2014starexec}
supply pinned corpora, labels, and the solver ecosystem's measured
state of the art — they rank solvers on fixed benchmarks. Saturation
inverts the roles: the benchmark is fixed and the \emph{instrument
grows} until the benchmark is exhausted, and the residue — the
terminal board, exported as challenge bundles — is a
machine-generated demand side for the next competition's entrants.

\paragraph{Relative completeness}
The framing of F4 is deliberately Cook's~\citep{cook1978soundness}:
completeness relative to an oracle for what cannot be internalized —
here, the candidate universe, builder competence, and the domain's
supply, each named as an assumption and each measured by the
platform's own books rather than assumed silently.
